UPRER–immunity axis acts as physiological food evaluation system that promotes aversion behavior in sensing low-quality food

Version 1

Abstract

Summary

To survive in challenging environments, animals must develop a system to assess food quality and adjust their feeding behavior accordingly. However, the mechanisms that regulate this chronic physiological food evaluation system, which monitors specific nutrients from ingested food and influences food-response behavior, are still not fully understood. Here, we established a low-quality food evaluation assay system and found that heat-killed E. coli (HK-E. coli), a low sugar food, triggers cellular UPRER and immune response. This encourages animals to avoid low-quality food. The physiological system for evaluating low-quality food depends on the UPRER (IRE-1/XBP-1) - Innate immunity (PMK-1/p38 MAPK) axis, particularly its neuronal function, which subsequently regulates feeding behaviors. Moreover, animals can adapt to a low-quality food environment through sugar supplementation, which inhibits the UPRER-PMK-1 regulated stress response by increasing vitamin C biosynthesis. This study reveals the role of the cellular stress response pathway as physiological food evaluation system for assessing nutritional deficiencies in food, thereby enhancing survival in nature environments.

Introduction

Food is essential for the survival, growth and fitness of all animals. To adapt to fluctuating environments with a wide range of food sources, animals have developed a food evaluation system. This system enables them to identify nutrient-rich food and avoids low-quality or toxic food, thereby maximizing their survival prospects 1–3. Various sensory neuron evaluation systems in animals has evolved to evaluate food quality through vision 4,5, olfactory 3,6–12 and gustatory senses 13–15. Besides these sensory systems that facilitate quick feeding decisions, animals may also initiate cellular stress response programs to detect nutrition/toxin and trigger food response behaviors 16,17. This could be one of physiological food quality evaluation systems that monitor the nutritional status of consumed food. However, the signaling events in cellular stress responses involved in evaluating of specific nutrients and the mechanisms that connect these signaling activities to food behaviors are largely unexplored. More specifically, while cellular stress response through UPRER 18 and PMK-1/p38 MAPK 19 dependent immunity in response to pathogens have been extensively studied, the functions of these cellular stress response in sensing and evaluating specific nutrients from food remain unclear.

Vitamin C is an essential micronutrient that cannot be synthesized by humans due to the loss of a key enzyme in the biosynthetic pathway 20. Animals obtain Vitamin C from their diet and possibly also from gut microbes 21. Vitamin C is an important physiological antioxidant and a cofactor for a family of biosynthetic and gene regulatory monooxygenase and dioxygenase enzymes. It is also required for the biosynthesis of collagen, L-carnitine, and certain neurotransmitters 22,23. Vitamin C has been associated with various human diseases including scurvy, immune defect and cardiovascular disease 20. Therefore, food evaluation systems in animals for sensing Vitamin C could be critical for their fitness in nature. However, the role of the cellular stress response pathway as food evaluation system for sensing and evaluating Vitamin C remains unexplored.

In this study, using the low-quality food evaluation assay system we established 24, we uncovered the mechanism of cellular stress responses pathway as physiological food evaluation system in evaluating D-Glucose deficiency in food and downstream Vitamin C content in animals through the UPRER (IRE-1/XBP-1) - Innate immunity (PMK-1 p38 MAPK) axis. This mechanism promotes animals to leave low-quality food and is critical for their survival in nature environments.

Results

Low-quality food induces stress response in animals

Our previous studies have shown that Heat-killed E. coli (HK-E. coli), which lacks certain molecules, is considered a low-quality food that is unable to support animal growth 24,25. Morevoer, through metabolic-seq analysis, we identified significant changes in a large numbers of metabolites (Figure S1A, Table S1), including lipids (Figure S1B, Table S1), amino acid and their metabolites (Figure S1C, Table S1), as well as coenzymes and vitamins (Figure S1D, Table S1). Interestingly, we observed a significant decrease in carboxylic acids and their derivatives (Figure S1E, Table S1) in E. coli after being heat-killed (Figure S1F, Table S1). This suggests that HK-E. coli is nutritionally deficient for C. elegnas when compared to normal E. coli food.

Next, we conducted two behavior assays to facilitate the analysis of the food evaluation process in animals by seeding L1 animals in assay plates (Figure 1A and 1B). In the avoidance assay, wild-type animals avoided the HK-E. coli food (Figure 1A). Interestingly, in the food choice assay, animals initially showed no preference between the two types of food (1-2h), but eventually exhibited a preference for high-quality food (Live E. coli) up until the17h mark (Figure 1B, Figure S1H). This suggests that worms depart from the HK-E. coli after recognizing it as low-quality food source through ingestion.

The stress response is induced in animals fed low-quality food, HK-E. coli.

(A) Schematic drawing and quantitative data of the food aversion assay. Circles indicate the food spot for live (yellow) and HK-OP50 (red) bacteria, respectively. The animals were scored 16-17 hours after L1 worms were placed on the food spot. Data are represented as mean ±SD from three independent experiments, 79-129 animals/assay.

(B) Schematic method and quantitative data of the food selection assay. Live (yellow), heat-killed (red) E. coli and LB as the buffer for E. coli were placed on indicated position. Synchronized L1 worms were place in the center spot. The selection index was calculated at the indicated time. Data are represented as mean ±SD from eight independent experiments, 123-792 animals/assay.

(C) GO enrichment analysis of up-regulated genes in animals fed with HK-E. coli vs live E. coli.

(D) Venn diagram showing numbers of UPRER target genes and up-regulated genes in animals fed HK-E. coli, and their overlap.

(E) Venn diagram showing numbers of induction genes by four pathogenic bacteria and HK-E. coli induced genes, and their overlap. The gene expression data was extracted from published data of animals’ infection with Pseudomonas aeruginosa (PA) 26, Bacillus thuringiensis (Bthu) 27, Staphylococcus aureus (Saur) 27, and Serratia marcescens (Smar) 27.

(F-G) GFP fluorescence images and bar graph showing that Phsp-4::GFP (F) and Pirg-5::GFP (G) were induced in animals fed wiht HK-E. coli. n is the number of worms scored from at least three independent experiments. Data are represented as mean ± SD.

For all panels, Scale bar shows on indicated figures, 50 μm. * p<0.05, ** p<0.01, *** p < 0.001, **** p<0.0001, ns: no significant difference. Precise P values are provided in Raw Data.

In order to investigate the pathways in animals that response to HK-E. coli, we performed transcriptomics analysis on worms that were cultured with both HK-E. coli and Live E. coli. Gene-expression profiling revealed that stress response genes, including those related to biotic stimulus, immune response and response to stress, are up-regulated in animals fed with HK-E. coli (Figure 1C, Table S2). Among these up-regulated genes, we identified 11 out of 62 of UPRER target genes (Figure 1D, Figure S1G and Table S2). Additionally, about 50%-80% of up-regulated genes overlap with genes responding pathogenic bacteria 26,27 (Figure 1E, Table S2), suggesting that UPRER and immune pathways may responded to low-quality food. Consistent with the results of the RNA sequencing (RNA-seq) analysis, the UPRER reporter (Phsp-4::GFP) and immunity reporter (Pirg-5::GFP) were strongly induced under HK-E. coli feeding condition (Figure 1F and 1G). Moreover, UPRMt reporter (Phsp-6::GFP) was weakly induced under HK-E. coli feeding condition (Figure S1I), and starved worm did not induce UPRER and immunity (Figure S1J and S1K).

Together, these findings suggest that low-quality food (HK-E. coli) triggers a stress response pathway in animals, including UPRER and innate immune. This implies that animals may assess the quality of food through UPRER and innate immune pathway.

Animals evaluate food quality through UPRER-immune depended physiological food quality evaluation system

To determine whether the UPRER and innate immune pathways play a role in evaluating low-quality food, we first examined whether the activation of the UPRER by HK-E. coli was dependent on the known signaling components of the UPRER branches, including IRE/XBP-1, PERK/ATF-4 and ATF-6 28,29. We observed no difference in Phsp-4::GFP induction with atf-4 (Figure S2A and S2J) or atf-6 (Figure S2B and S2J) RNAi-mediated knockdown in animals fed with HK-E. coli. However, knockdown of ire-1/xbp-1 or mutation of xbp-1 reduced GFP fluorescence (Figure 2A, 2B and S2J). Among the 11 deferentially expressed UPRER target genes in animals fed with HK-E. coli from RNA-seq (Figure 1D), 64% of the genes are IRE-mediated genes (Figure S1G, Table S2). The mRNA level of IRE-1-mediated splicing of xbp-1 is also induced in animals fed with HK-E. coli (Figure S2C). These data suggest that activation of the UPRER by low-quality food depends on the IRE-1/XBP-1.

Animals evaluate food quality through UPRER (ire-1/xbp-1) - Innate immunity (pmk-1 MAPK) axis.

(A-B) GFP fluorescence images and bar graph showing that HK-E. coli induced Phsp-4::GFP was decreased in animals with ire-1 or xbp-1 RNAi treatment (A) or xbp-1 mutation (B). n is the number of worms scored from at least three independent experiments. Data are represented as mean ± SD.

(C) Food aversion assay showing that xbp-1 mutation eliminated the discrimination against HK-E. coli. However, this effect is rescued by expressing xbp-1 in neurons rather than intestine. Data are represented as mean ±SD from five independent experiments, 156-763 animals/assay.

(D) GFP fluorescence images and bar graph showing that HK-E. coli induced Pirg-5::GFP was decreased in animals with pmk-1 RNAi treatment. n is the number of worms scored from at least three independent experiments. Data are represented as mean ± SD.

(E) Western blot images showing the level of p-PMK-1 in L1 animals (Wild-type N2 and xpb-1 mutant) fed with OP50 or HK-OP50 for 4 h. The level of p-PMK-1 is induced in animals fed HK-OP50.

(F) Food aversion assay showing that pmk-1 mutation eliminated the discrimination against HK-E. coli. However, this effect is rescued by expressing pmk-1 in neurons rather than intestine. Data are represented as mean ±SD from five independent experiments, 168-492 animals/assay.

(G) Food aversion assay in wild-type, xbp-1, pmk-1 and double mutant. Data are represented as mean ±SD from five independent experiments, 259-490 animals/assay..

For all panels, Scale bar shows on indicated figures, 50 μm. * p<0.05, ** p<0.01, *** p < 0.001, **** p<0.0001, ns: no significant difference. Precise P values are provided in Raw Data.

To further analyze whether XBP-1-dependent UPRER activation is critical for animals to leave low-quality food, we tested food avoidance behavior using xbp-1 mutant. The results show that xbp-1 mutants had a significantly decreased likelihood of leaving of HK-E. coli, which was rescued by expressing xbp-1 in neuron rather than intestine (Figure 2C). This indicats that XBP-1-dependent UPRER activation in neuron is critical for animals to evaluate low-quality food.

We then investigated which innate immune pathway is involved in evaluating low-quality food. First, we analyzed HK-E. coli induced genes from RNA-seq. Among these up-regulated genes, 82 out of 409 of PMK-1-dependent genes 30 were identified (Figure S2D, Table S2). Second, we confirmed the induction of several well-known PMK-1 target genes in RNA-seq data31 (Figure S2E) and reporter analysis (Figure S2F and S2G). Moreover, the induction of Pirg-5::GFP was abolished in pmk-1 knockdown animals fed with HK-E. coli (Figure 2D and S2K). Third, we found that the phosphorylated PMK-1 (p-PMK-1) level was prominently increased in wild-type N2 animals fed HK-E. coli compared to feeding E. coli OP50 (Figure 2E). Finally, pmk-1 mutants had a decreased likelihood of leaving of HK-E. coli, which was rescued by expressing pmk-1 in neurons rather than intestine (Figure 2F). These data suggest that PMK-1 regulated immune pathway evaluates low-quality food, especially the neuronal PMK-1 has a critical function for food quality response.

XBP-1 and PMK-1 are in the same pathway for evaluating food quality

Next, we explored the connection between UPRER (IRE-1/XBP-1) and innate immunity (PMK-1 p38 MAPK) in food quality evaluation. We found that Pirg-5::GFP induction (Figure S2H and S2K) and PMK-1 activation (Figure 2E) were decreased in animals with xbp-1 mutation or knockdown when fed with HK-E. coli, suggesting that XBP-1 could regulate PMK-1 under this condition. Additionally, Phsp-4::GFP induction under HK-E. coli was not affected in animals with pmk-1 RNAi (Figure S2I), indicating that XBP-1-dependent UPRER activation is not regulated by PMK-1. Finally, we constructed a double mutant of xbp-1 and pmk-1 and found that the food avoidance phenotype of the double mutant was similar to the pmk-1 mutant (Figure 2G), indicating that PMK-1 is downstream of XPB-1 in responding to low-quality food.

Sugar deficiency in HK-E. coli food induces stress response and avoidance behavior in animals

We then investigated which nutrients/metabolites are sensed by animals through the XBP-1-PMK-1 axis for food quality evaluation. First, we hypothesized that the nutrition status is improved in E. coli mutant (HK-treatment), which could inhibit URPER and immune response in animals (Figure S3A). We established a system for screening the E. coli mutant keio library (Figure S3A), and identified 20 E. coli mutants that did not induce hsp-4::GFP through the UPRER reporter (irg-5::GFP) after three rounds of screening (Table S3). From these 20 E. coli mutant, we identified 9 E. coli mutants that did not induce irg-5::GFP through the immunity reporter (irg-5::GFP) screening (Figure S3B, S3C, Table S3). Animals fed HK-yfbR, which catalyzes carbohydrate derivative metabolic process32, had a decreased ability to leave food (Figure 3A, Table S3), indicating that HK-yfbR may be a higher quality food for animals compared to HK-K12.

HK-E. coli is low sugar food, which induce stress response and avoidance behavior in animals.

(A) Food aversion assay showing that wild-type animals eliminated the discrimination against HK-E. coli when yfbR is mutated in E. coli. Data are represented as mean ±SD four independent experiments, 251-490 animals/assay.

(B) Heat map showing the thirteen differential metabolites from HK-K12, HK-yfbR, and K12 in 4 independent experiments. Color indicates the relative level of each metabolite.

(C-D) The bar graph showing that HK-E. coli induced Phsp-4::GFP was decreased in animals with lactose (C) or D-(+)-sucrose (D) supplementation. n is the number of worms scored from at least three independent experiments. Data are represented as mean ± SD.

(E-F) GFP fluorescence images and bar graph showing that HK-E. coli induced Phsp-4::GFP (E) and Pirg-5::GFP (F) were decreased in animals with D-(+)-Glucose (D-Glc) supplementation. n is the number of worms scored from at least three independent experiments. Data are represented as mean ± SD.

(G) Western blot images showing the level of p-PMK-1 in L1 animals fed HK-E. coli with or without D-(+)-Glucose (D-Glc) supplementation for 4 h. The level of p-PMK-1 is decreased in animals fed HK-OP50+D-Glc.

(H) Food aversion assay showing that wild-type animals eliminated the discrimination against HK-E. coli with D-Glc supplementation. Data are represented as mean ±SD six independent experiments, 190-492 animals/assay.

For all panels, Scale bar shows on indicated figures, 50 μm. * p<0.05, ** p<0.01, *** p < 0.001, **** p<0.0001, ns: no significant difference. Precise P values are provided in Raw Data.

Secondly, we preformed a metabolomics analysis of different quality food (HK-K12, HK-yfbR and Live-K12). We found that the level of 13 metabolites were similar between HK-yfbR and Live-K12, but significantly changed in HK-K12 (Figure 3B, Figure S3D, and Table S1). We also found that genes involved in glycolysis/gluconeogenesis were up-regulated in animals fed with HK-E. coli (Figure S3E), suggesting that glycolysis/gluconeogenesis metabolism is disordered in animals fed with HK-E. coli, which may result from changes in sugar/carbohydrate intake. The carbohydrates (D-trehalose, Lactose, and D-(+)-sucrose) were also decreased in HK-E. coli (Figure 3B), suggesting that carbohydrate deficiency may induce stress response and avoidance behavior in animals.

Thirdly, to determine which carbohydrate inhibits stress response in animals, we supplemented each metabolite to HK-E. coli and found that only Lactose, and D-(+)-sucrose inhibited HK-E. coli induced UPRER (Figure 3C and 3D, Figure S3F). Moreover, we found from our metabolomic data that the sugar level, including Lactose, and D-(+)-sucrose, and D-(+)-Glucose, was also decreased in HK-E. coli (Figure 3B, Table S1). Since Lactose and D-(+)-sucrose are hydrolyzed to produce glucose33,34, we wondered whether glucose also inhibits the stress response in animals. We found that D-(+)-Glucose supplementation also inhibited HK-E. coli induced UPRER (Figure 3E), immune response (Figure 3F, 3G and Figure S3G) and avoidance (Figure 3H). Moreover, sugar supplementation did not affect UPRER and immunity in normal food (OP50) or starved condition (NGM) (Figure S3H and S3I).

Together, these findings suggest that sugar/carbohydrates deficiency in HK-E. coli induces a stress response and avoidance behavior in animals, which can be inhibited by D-(+)-Glucose supplementation. This implies that animals may evaluate the sugars deficiency through the activation of UPRER and immune responses.

Animals could overcome a low-quality food environment by sugar supplementation through vitamin C biosynthesis

We discovered that D-(+)-Glucose supplementation inhibited HK-E. coli induced UPRER (Figure 3E), immune response (Figure 3F, 3G and Figure S3G) and avoidance (Figure 3H). Simultaneously, vitamin C (VC), which is synthesized by glucuronate pathway using D-Glucose 35,36 (Figure 4A), was found contribute to neuroprotective37,38, immune defense39,40, and inhibits inflammatory and ER stress41,42. This led us to question whether the Vitamin C biosynthesis pathway is involved in evaluating low-quality food by using D-Glucose.

Vitamin C biosynthesis pathway is critically involved in evaluating sugar in the food

(A) Cartoon illustration of a simplified, Vitamin C biosynthesis pathway in C. elegans. The relevant coding genes of enzymes was labeled with red.

(B) The level of total L-ascorbic acid (TAA) in animals fed with Live-K12, HK-K12, or HK-yfbR. Data are represented as mean ±SD from six independent experiments.

(C) The level of total L-ascorbic acid (TAA) in animals (control or knockdown of Vitamin C biosynthesis genes) fed with HK-E. coli with or without D-glucuronate (D-GlcA) supplementation. Data are represented as mean ±SD from eight independent experiments.

(D-E) GFP fluorescence images and bar graph showing that HK-E. coli induced Phsp-4::GFP (C) and Pirg-5::GFP (D) were decreased in animals with D-GlcA or Vitamin C supplementation. n is the number of worms scored from at least three independent experiments. Data are represented as mean ± SD.

(F-G) Western blot images showing the level of p-PMK-1 in L1 animals fed with HK-E. coli with D-GlcA or Vitamin C supplementation for 4 h. The level of p-PMK-1 is decreased in animals with D-GlcA (F) or Vitamin C (G) supplementation.

(H) Food aversion assay showing that wild-type animals eliminated the discrimination against HK-E. coli with D-GlcA or Vitamin C supplementation. N is number of independent experiments. Data are represented as mean ±SD from four independent experiments, 153-292 animals/assay.

(I-K) The bar graph showing that suppression of HK-E. coli induced Phsp-4::GFP (I), Pirg-5::GFP (J) and food avoidance (K) by D-GlcA supplementation was abolished in animals with RNAi of VC biosynthesis genes, which was not affect by Vitamin C supplementation. n is the number of worms scored from at least three independent experiments and Data are represented as mean ±SD(I-J). Data are represented as mean ±SD from five independent experiments, 252-537 animals/assay (K).

For all panels, Scale bar shows on indicated figures, 50 μm. * p<0.05, ** p<0.01, *** p < 0.001, **** p<0.0001, ns: no significant difference. Precise P values are provided in Raw Data.

Firstly, we observed an increase in the vitamin C level in animals when fed with HK-yfbR (Figure 4B), a high carbohydrate food compared to HK-E. coli (Figure 3B and Figure S1E). However, the VC level in bacteria is same (Figure S4A). The VC level also increased when D-Glucose (Figure S4B) or D-glucuronate (D-GlcA) was added to HK-E. coli (Figure 4C), which was abolished by knocking down VC biosynthesis genes (Figure 4C and S4B). This suggests that additional of sugar supplementation promotes VC synthesis in animals fed with HK-E. coli.

Secondly, we hypothesized that animals could overcome a low-quality food (HK-E. coli) environment by inhibiting the stress response through increasing vitamin C biosynthesis. We found that VC or D-glucuronate (D-GlcA) supplementation inhibited HK-E. coli induced UPRER (Figure 4D), immune response including irg-5/sysm-1 reporter expression (Figure 4E and Figure S4C) and p-PMK-1 (Figure 4F and 4G), as well as food avoidance (Figure 4H).

Finally, we asked whether inhibition of stress response and avoidance by sugar supplementation depends on the vitamin C biosynthesis pathway. We found that suppression of HK-E. coli induced UPRER (Figure 4I), immune response (Figure 4J and Figure S4D) and food avoidance (Figure 4K) by D-GlcA/sugar supplementation was abolished in animals with RNAi of VC biosynthesis genes. Food selection behavior assays showed that D-GlcA/sugar supplementation increased the preference for heat-killed bacteria, which was also suppressed by knocking down VC biosynthesis genes (Figure S4E). However, VC supplementation still suppressed the UPRER (Figure 4I), immune response (Figure 4J and Figure S4D) and food avoidance (Figure 4K), and increased the food preference (Figure S4E) in animals with or without RNAi of VC biosynthesis genes. This suggests that VC, as the final metabolite synthesized from D-Glucose, is critical for low-quality food response in animals.

Together, these data indicate that the Vitamin C biosynthesis pathway is critical for evaluating whether food is of higher quality and can be eaten by animals. It also suggests that animals could improve their VC levels to adapt to bad food environment.

Animals evaluate sugar and vitamin C through neuronal XBP-1 and PMK-1

As D-GlcA/sugar and VC supplementation suppressed HK-E. coli induced UPRER, immune response and food avoidance behavior, we investigated whether animals evaluate sugar and VC through XBP and PMK-1 dependent pathways. We performed a food selection behavior assay by adding D-Glc, D-GlcA or VC to the NGM, E. coli and HK-E. coli (Figure 5A). The food selection behavior assays revealed that supplementation with D-Glc, D-GlcA, or VC inhibits the animals’ choice of sugar or VC on E. coli-OP50 feeding conditons (Figure S4F). This suggests that supplementation with D-Glc, D-GlcA, or VC may alter the metabolites of live bacteria, leading to avoidance by the animals. There was no preference observed on NGM (no food condition) supplementation with D-Glc and VC (Figure S4G), indicating that the intake of sugar or VC alone does not influence animal preference. However, alone D-GlcA could influence worm physiology which induces preference change (Figure S4G). Interestingly, D-Glc and D-GlcA (Figure 5B and 5C) or VC (Figure 5D) supplementation increased the preference for heat-killed bacteria, which was suppressed in xbp-1 or pmk-1 mutant animals. However, this preference was also rescued in xbp-1 or pmk-1 mutant animals by expressing XBP-1 or PMK-1 in neurons rather than intestine (Figure 5B-D), indicating that neuronal XBP-1 and PMK-1 are critical for physiological food elevation system for monitoring the level of sugar and VC under low-quality food condition.

Animals evaluate sugar and vitamin C through neuronal XBP-1 and PMK-1

(A) Schematic method of the food selection assay. Heat-killed E. coli (red) and HK-E. coli with chemical supplementation (green) were placed on indicated position. Synchronized L1 worms were then place in plate. After 16-17hs, the selection index was calculated.

(B-D) Food selection assay showing that xbp-1 or pmk-1 mutation eliminated the preference of HK-E. coli with D-Glc (B), D-GlcA (C) or Vitamin C (D) supplementation, which was rescued in xbp-1 or pmk-1 mutant animals by expressing XBP-1 or PMK-1 in neurons rather than intestine.

Data are represented as mean ±SD from five independent experiments, 68-647 animals/assay (B).

Data are represented as mean ±SD from six independent experiments, 83-701 animals/assay (C).

Data are represented as mean ±SD from six independent experiments, 67-1035 animals/assay (D).

For all panels, Scale bar shows on indicated figures, 50 μm. * p<0.05, ** p<0.01, *** p < 0.001, **** p<0.0001, ns: no significant difference. Precise P values are provided in Raw Data.

Schematic model of physiological food evaluation system in evaluating/sensing sugar and vitamin C through UPRER (IRE-1/XBP-1) - Innate immunity (PMK-1/p38 MAPK) axis.

Vitamin C level is low in animals fed low sugar food, HK-E. coli. Sugar and Vitamin C deficiency activate cellular UPRER and immune response, which promote animals to leave low-quality food and seek better food for survival. This cellular stress regulated physiological food evaluation system depends UPRER (IRE-1/XBP-1) - Innate immunity (PMK-1/p38 MAPK) axis in neuron.

Discussion

To better survive, animals must evolve a system to recognize and evaluate the quality of their food. This includes the sensory neuron evaluation system for immediate response and feeding decision 2, as well as physiological food evaluation system for chronic response to ingested food. In our previous study, we discovered that the TORC1-ELT-2 pathway, acting as master regulators in intestine, evaluates vitamin B2 deficiency in low-quality food (HK-E. coli) and regulates gut digestive activity to impact animal’s food behavior 24. To further identified the mechanism by which animals evaluate low-quality food (HK-E. coli), we performed metabolomics and transcriptomics analyses to identify specific nutrition deficiencies in low-quality food and the cellular response pathways that are involved in food evaluation pathway. This study identified a physiological food evaluation mechanism by which animals detect sugar (in food) and vitamin C (in animals) deficiency through UPRER (IRE-1/XBP-1) - Innate immunity (PMK-1/p38 MAPK) regulated cellular stress response program in neurons that dictates food avoidance and selection behaviors.

Unlike the sensory neuron evaluation system, which permits rapid feeding decisions through smell and taste, the cellular stress response as physiological food evaluation system describe here requires a slow and multi-step signal transduction process after the ingestion of food. The disruption of cellular homeostasis by ingested of low-quality or toxic food can activate stress response mechanisms that both increase the cellular ability to withstand and adapt to this disruption of homeostasis and promote behavioral strategies to avoid these conditions and lessen their impact on the organism. These cellular stress response mechanisms include heat shock response, unfolded protein response, oxidative stress response 43. Therefore, this slow physiological food evaluation system is an evolutionary adaptation mechanism for detecting nutrition deficiencies in food that was not detected by quick sensory neuronal system.

One of cellular stress response, UPRER, is activated by stresses such as infection and nutrition deficiency that affect homeostasis in the endoplasmic reticulum (ER) 28. The activation of the UPRER by expression of xbp-1s in the nervous system has recently been shown to promote changes in feeding and foraging behavior 44. The p38 PMK-1 pathway regulates the expression of secreted innate immune effectors and is required for survival during infection 45. Moreover, UPRER and innate immunity increased by disrupting in vivo state would induce worm aversion from food lawn46–49. Thus, these two cellular stress response pathways are critical for animals’ survival in changing environment. However, whether UPRER (IRE-1/XBP-1) - Innate immunity (PMK-1/p38 MAPK) evaluates food quality under physiological conditions is largely unexplored. Our study presents evidence that low-quality food (HK-E. coli) activate UPRER and PMK-1, and promotes animals to leave food. It is independent of avoidance behavior of hyperactivation of kgb-1/MAPK pathway49 (Figure S4H). Mover, our study also shows that neuronal UPRER and PMK-1 are critical for evaluating low-quality food. Therefore, our study established a novel physiological food evaluation system by activating the cellular stress response program with UPRER (IRE-1/XBP-1) - Innate immunity (PMK-1/p38 MAPK).

Collectively, this study uncovers the unexpected function of UPRER (IRE-1/XBP-1) - Innate immunity (PMK-1/p38 MAPK) as a physiological food evaluation system for evaluating and sensing food quality in animals. It also highlights the utility of the HK-E. coli (low-quality food) - C. elegans interaction as a means to dissect the mechanism of food evaluation system in assessing food. Most importantly, it reveals that animals are capable of altering their nutrient (Vitamin C) levels through in vivo synthesis or food intake to adapt to a poor food environment when better food choices are not available.

Author Contributions

P. L designed, performed experiments, analyzed data. X. L constructed all transgenic animals. B.Q. designed research, supervised this study, and wrote the paper with inputs from P. L.

eLife Assessment

This important study presents interesting findings that suggest that the UPR and immune regulators can act as evaluators of nutritional quality in C. elegans. Convincing evidence expands our understanding both of physiological food evaluation systems and of the known roles of stress response pathways in organismal physiology. However, there is limited mechanistic exploration in the study, and in some cases, the effect size is small and statistical significance questionable.

Reviewer #1 (Public Review):

Summary:

This manuscript by Liu et al explores the role of the UPR and immune regulators in the evaluation of nutritional quality in C. elegans. They identify neuronal UPR activation and the MAPK PMK-1 as key responders to low food quality. In particular, the data suggest that these pathways are activated by low levels of vitamin C synthesis that result from the low sugar levels present in heat-killed E. coli.

Strengths:

The results are intriguing and expand our understanding both of physiological food evaluation systems, and of the known roles of stress response pathways in organismal physiology. The authors use a range of techniques, encompassing imaging, metabolomic analysis, gene expression analysis, and behavioural assays, to support their claims.

Weaknesses:

There is limited mechanistic analysis in the study. In particular, how does low vitamin C trigger UPR activation? This is an intriguing finding that, if followed up, could potentially reveal a novel mechanism of UPR activation. In addition, how is the activation of the PMK-1 pathway driven by/coordinated with UPR activation? The data in some figures is not as convincing as it could be: the magnitude of the effect size is small in the supplementation experiments, and the statistical tests used are not always appropriate to enable multiple comparisons.

Reviewer #2 (Public Review):

Summary:

In this work, the authors aim to better understand how C. elegans detects and responds to heat-killed (HK) E. coli, a low-quality food. They find that HK food activates two canonical stress pathways, ER-UPR, and innate immunity, in the nervous system to promote food aversion. Through the creative use of E. coli genetics and metabolomics, the authors provide evidence that the altered carbohydrate content of HK food is the trigger for the activation of these stress responses and that supplementation of HK food with sugars (or their biosynthetic product, vitamin C), reduces stress pathway induction and food avoidance. This work makes a valuable addition to the literature on metabolite detection as a mechanism for the evaluation of nutritional value; it also provides some new insight into the physiologically relevant roles of well-known stress pathways in modulating behavior.

Strengths:

-The work addresses an important question by focusing on understanding how the nervous system evaluates food quality and couples this with behavioral change.

-The work takes full advantage of the tools available in this powerful system and builds on extensive previous studies on feeding behavior and stress responses in C. elegans.

-Creative use of E. coli genetics and metabolite profiling enabled the identification of carbohydrate metabolism as a candidate source of food-quality signals.

-For the most part, the studies are rigorous and logically designed, providing good support for the authors' model.

Weaknesses:

-It is not clear how the mechanism identified here is connected to previously described, related processes. In particular, it is not clear whether this mechanism has a role in the detection of other low-quality foods. Further, the specificity of the ability of sugar/vitamin C to suppress stress pathway induction is unclear (i.e., does sugar/vitamin C have any effect on the activation of these pathways through other means?). Additionally, the relationship of this pathway to the vitamin B2-sensing mechanism previously described by the senior author is unclear. These issues do not weaken confidence in the authors' conclusions, but they do reduce the potential significance of the work.

-The authors claim that the induction of the innate immune pathway reporter irg-5::GFP is "abolished" in pmk-1(RNAi) animals, but Figure S2K seems to show a clear GFP signal when these animals are fed HK-OP50. Similarly, the claim that feeding WT animals HK-OP50 enriches phospho-PMK-1 levels (Fig 2E) is unconvincing - only one western blot is shown, with no quantification, and there is a smear in the critical first lane.

-The rationales for some of the paper's hypotheses could be improved. For example, the rationale for screening the E. coli mutant library is that some mutants, when heat-killed, may be missing a metabolite that induces the ER-UPR. A more straightforward hypothesis might be that some mutant E. coli strains aberrantly induce the ER-UPR when *not* heat-killed, because they are missing a metabolite that prevents stress pathway induction. This is not in itself a major concern, but it would be useful for the authors to provide a rationale for their hypothesis.

-The authors do not provide any explanation for some unexpected results from the E. coli screen. Earlier in the paper, the authors found that innate immune signaling is downstream of ER-UPR activation. However, of the 20 E. coli mutants that, when heat-killed, "did not induce... the UPR-ER reporter," 9 of them still activate the innate immune response. This seems at odds with the authors' simple model since it suggests that low-quality food can induce innate immune signaling independently of the ER-UPR. Further, only one of the 9 has an effect on behavior, even though failure to activate the innate immune pathway might be expected to lead to a behavioral defect in all of these.

-In a number of places, the writing style can make the authors' arguments difficult to follow.

-Some of the effect sizes observed by the authors are exceedingly small (e.g, the suppression of hsp-4::gfp induction by sugar supplementation in Figs 3C-E), raising some concern about the biological significance of the effect.

-In some cases, there is a discrepancy between the fluorescence images and their quantitation (e.g., Figure 3E, where the effect of glucose on GFP fluorescence seems much stronger in the image than in the graph).

Reviewer #3 (Public Review):

Summary:

Animals can evaluate food quality in many ways. In contrast to the rapid sensory evaluation with smell and taste, the mechanism of slow nutrient sensation and its impact on food choice is unexplored. The authors utilize C. elegans larvae and their bacterial food as an elegant model to tackle this question and reveal the detailed molecular mechanism to avoid nutrient-poor foods.

Strengths:

The strength of this study is that they identified the molecular identities of the critical players in bacterial food and C. elegans using unbiased approaches, namely metabolome analysis, E. coli mutant screening, and RNA sequencing. Furthermore, they strengthen their findings by thorough experiments combining multiple methods such as genetics, fluorescent reporter analysis, and Western blot.

Weaknesses:

The major caveat of this study is the reporter genes. The transcriptional reporters were used to monitor the UPRER and immune responses in the intestine of C. elegans. However, their tissue-specific rescue experiments suggest that the genes in the UPRER and immune response function in the neurons. Thus, we should carefully interpret the results of the reporter genes.

Overall, this work provides convincing data to support their model. In the C. elegans field, the behaviors of larvae are not well studied compared to adults. This work will pose an interesting question about the difference between larvae and adults in nutrition sensing in C. elegans and provide a framework and candidate molecules to be studied in other organisms.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public Review):

Summary:

This manuscript by Liu et al explores the role of the UPR and immune regulators in the evaluation of nutritional quality in C. elegans. They identify neuronal UPR activation and the MAPK PMK-1 as key responders to low food quality. In particular, the data suggest that these pathways are activated by low levels of vitamin C synthesis that result from the low sugar levels present in heat-killed E. coli.

Strengths:

The results are intriguing and expand our understanding both of physiological food evaluation systems, and of the known roles of stress response pathways in organismal physiology. The authors use a range of techniques, encompassing imaging, metabolomic analysis, gene expression analysis, and behavioural assays, to support their claims.

Thank you for your thorough review and acknowledgment of the strengths of our study.

Weaknesses:

There is limited mechanistic analysis in the study. In particular, how does low vitamin C trigger UPR activation? This is an intriguing finding that, if followed up, could potentially reveal a novel mechanism of UPR activation. In addition, how is the activation of the PMK-1 pathway driven by/coordinated with UPR activation? The data in some figures is not as convincing as it could be: the magnitude of the effect size is small in the supplementation experiments, and the statistical tests used are not always appropriate to enable multiple comparisons.

(1) There is limited mechanistic analysis in the study. In particular, how does low vitamin C trigger UPR activation? This is an intriguing finding that, if followed up, could potentially reveal a novel mechanism of UPR activation.

Thank you for highlighting the need for further mechanistic analysis in our study. We appreciate the opportunity to clarify the process by which low vitamin C triggers UPR activation.

Our investigation revealed that the vitamin C content in heat-killed E. coli (HK-E. coli) is comparable to that of live E. coli or HK-yfbR mutant E. coli (Figure 4-figure supplement 1A), indicating that the induction of unfolded protein response (UPR) in C. elegans by HK-E. coli is not solely attributed to low vitamin C levels but rather involves other unidentified factors.

Through metabolomic analysis, we observed significant decreases in sugar levels, including lactose, D-(+)-sucrose, and D-(+)-glucose, in HK-E. coli (Figure 3B, Table S1). Notably, supplementing D-(+)-glucose effectively inhibited UPRER, immune response, and avoidance behavior induced by HK-E. coli (Figure 3E-H). These findings suggest that the deficiency in sugars in HK-E. coli triggers a stress response and avoidance behavior in animals, which can be alleviated by D-(+)-glucose supplementation.

Furthermore, when comparing heat-killed E. coli mutant yfbR (HK-yfbR) to HK-E. coli, we observed significantly higher sugar levels, including lactose and D-(+)-sucrose, in HK-yfbR (Figure 3B). This was accompanied by reduced UPRER in animals feeding on HK-yfbR (Figure 3-figure supplement 1B), indicating that higher sugar levels may inhibit the induction of UPRER by low-quality food.

Considering that the synthesis of vitamin C (VC) occurs through the glucuronate pathway, utilizing D-glucose as a precursor 1, 2 (Figure 4A), we investigated whether the vitamin C biosynthesis pathway is involved in evaluating low-quality food using D-glucose. Contrary to our initial hypothesis, animals fed live E. coli did not exhibit higher glucose levels compared to those fed low-quality food (HK_-E. coli_). Our results indicate that animals maintain similar VC levels when fed ideal food (live E. coli) compared to low-quality food (HK-E. coli) (Figure 4B), suggesting that animals do not stimulate VC biosynthesis under favorable food conditions. However, supplementation of D-GlcA or E. coli-yfbR mutation in HK-E. coli significantly improved VC levels when animals were fed low-quality food (HK-OP50) (Figure 4B, 4C). Moreover, VC or D-glucuronate (D-GlcA) supplementation inhibited HK-E. coli-induced UPRER (Figure 4D), indicating that glucose boosts the animal's ability to adapt to unfavorable food environments by increasing VC levels, thereby inhibiting UPRER, but not under favorable food conditions.

These findings shed light on the complex interplay between vitamin C, sugar levels, and UPR activation, providing valuable insights into the mechanisms underlying food evaluation and stress response pathways in organisms.

Overall, we are grateful for the reviewer's constructive feedback, which motivates us to continue our efforts to understanding how the UPR response contributes to the complexities of food evaluation and behavioral responses in organisms.

(2) In addition, how is the activation of the PMK-1 pathway driven by/coordinated with UPR activation?

Thank you for your insightful inquiry. In our discussion section, we have addressed this question by integrating new data and discussion to provide insights into the coordination between PMK-1 pathway activation and UPR activation.

Previous studies have demonstrated that activating innate immunity, specifically the PMK-1 MAPK pathway, results in a reduction in translation3, as well as a shutdown of food digestion in animals4, likely aimed at reducing protein translation and cellular metabolism. To further investigate this relationship, we measured the translation level of animals fed with heat-killed E. coli (HK-E. coli) and found a significant reduction in total translation ability in these animals (Figure 5-figure supplement 1D). This observation suggests that activating innate immunity through the PMK-1 MAPK pathway may serve as a mechanism to slow down translation progress, thereby alleviating the pressure on the unfolded protein response (UPR) and preventing excessive UPRER activation.

By integrating these findings, we propose a model wherein activation of the PMK-1 pathway coordinates with UPR activation to regulate translation and cellular metabolism in response to low-quality food. This coordinated response likely serves to maintain cellular homeostasis and prevent detrimental effects associated with excessive UPRER activation.

These insights contribute to our understanding of the intricate interplay between innate immunity, cellular stress responses, and metabolic regulation in organisms facing nutritional challenges.

(3) The data in some figures is not as convincing as it could be: the magnitude of the effect size is small in the supplementation experiments, and the statistical tests used are not always appropriate to enable multiple comparisons.

We appreciate the reviewers' concerns regarding the data presentation and statistical analyses in some of our figures. In response to this feedback, we have made revisions to improve the robustness and clarity of our statistical methods.

All statistical analyses were conducted using GraphPad Prism 8.0 software. Specifically, a two-tailed unpaired t-test was employed for the statistical analysis of two groups of samples, while one-way or two-way ANOVA was utilized for the statistical analysis of more than two groups of samples. These adjustments ensure appropriate statistical comparisons and enhance the reliability of our findings.

Reviewer #2 (Public Review):

Summary:

In this work, the authors aim to better understand how C. elegans detects and responds to heat-killed (HK) E. coli, a low-quality food. They find that HK food activates two canonical stress pathways, ER-UPR, and innate immunity, in the nervous system to promote food aversion. Through the creative use of E. coli genetics and metabolomics, the authors provide evidence that the altered carbohydrate content of HK food is the trigger for the activation of these stress responses and that supplementation of HK food with sugars (or their biosynthetic product, vitamin C), reduces stress pathway induction and food avoidance. This work makes a valuable addition to the literature on metabolite detection as a mechanism for the evaluation of nutritional value; it also provides some new insight into the physiologically relevant roles of well-known stress pathways in modulating behavior.

Strengths:

-The work addresses an important question by focusing on understanding how the nervous system evaluates food quality and couples this with behavioral change. -The work takes full advantage of the tools available in this powerful system and builds on extensive previous studies on feeding behavior and stress responses in C. elegans.

-Creative use of E. coli genetics and metabolite profiling enabled the identification of carbohydrate metabolism as a candidate source of food-quality signals.

-For the most part, the studies are rigorous and logically designed, providing good support for the authors' model.

We deeply appreciate the reviewer's insightful assessment of our study's strengths.

Weaknesses:

-It is not clear how the mechanism identified here is connected to previously described, related processes. In particular, it is not clear whether this mechanism has a role in the detection of other low-quality foods. Further, the specificity of the ability of sugar/vitamin C to suppress stress pathway induction is unclear (i.e., does sugar/vitamin C have any effect on the activation of these pathways through other means?). Additionally, the relationship of this pathway to the vitamin B2-sensing mechanism previously described by the senior author is unclear. These issues do not weaken confidence in the authors' conclusions, but they do reduce the potential significance of the work.

(1) In particular, it is not clear whether this mechanism has a role in the detection of other low-quality foods.

Thank you for your valuable feedback. In response to your inquiry, we investigated whether the UPRER (IRE-1/XBP-1) - Innate immunity (PMK-1/p38 MAPK) axis is specific to evaluating low-quality food (HK-E. coli) or if it plays a broader role in food detection.

We conducted behavioral assays using N2, pmk-1, and xbp-1 mutant animals fed with normal E. coli food, inedible food (Saprophytic staphylococci)4, and pathogenic food (Pseudomonas aeruginosa-PA14)5. We found that N2, pmk-1, and xbp-1 mutant worms did not exhibit avoidance behavior when presented with normal food (OP50). However, both N2 and xbp-1 mutant worms were able to escape from inedible food (N2 was predominantly found on the border areas of the bacterial lawn and xbp-1 mutant worms on border and in), Saprophytic staphylococci, whereas pmk-1 mutant worms did not exhibit this avoidance behavior. Notably, N2 and xbp-1 mutant worms exhibited even more pronounced avoidance behavior when exposed to Pseudomonas aeruginosa, whereas pmk-1 mutant worms were more susceptible to infection by this pathogen (Figure 2-figure supplement 2C). These findings suggest that the UPR-Immunity pathway plays a crucial role in helping animals avoid low-quality food (HK-E. coli) by triggering an avoidance response. In contrast, the Innate immunity pathway, mediated by PMK-1/p38 MAPK, appears to play a key role in evaluating unfavorable food sources, such as HK-E. coli, Saprophytic staphylococci, and Pseudomonas aeruginosa, and helping animals avoid these environments.

(2) Further, the specificity of the ability of sugar/vitamin C to suppress stress pathway induction is unclear (i.e., does sugar/vitamin C have any effect on the activation of these pathways through other means?).

Thank you for your inquiry regarding the specificity of the ability of sugar/vitamin C to suppress stress pathway induction. We aimed to address this question by investigating whether high levels of VC inhibit other stress-induced UPRER pathways.

Previous studies have shown that both Tunicamycin6 and pathogenic bacteria, such as Pseudomonas aeruginosa-PA145, induce UPRER in C. elegans. In response to your query, we conducted experiments to examine whether VC supplementation inhibits UPRER induced by these stressors. Our findings indicate that VC supplementation does not inhibit UPRER induced by either Tunicamycin or PA14 (Author response image 1).

These results suggest that while sugar/vitamin C may suppress stress pathway induction in the context of low-quality food, its effects may not extend to other stressors that induce UPRER through different mechanisms. This insight helps clarify the specificity of sugar/vitamin C's role in modulating stress pathway activation, contributing to a better understanding of the broader regulatory networks involved in stress response in C. elegans.

VC supplementation does not inhibit Tunicamycin or PA14-induced UPRER.

(3) Additionally, the relationship of this pathway to the vitamin B2-sensing mechanism previously described by the senior author is unclear.

In response to your comment, we would like to clarify the relationship of our pathway to the previously described vitamin B2-sensing mechanism we found. Previous studies have demonstrated that heat-killed E. coli (HK-E. coli) serves as a low-quality food source incapable of supporting the growth of C. elegans larvae, whereas supplementation with vitamin B2 (VB2) can restore animal growth7

This study investigates the role of sugar deficiency in HK-E. coli, which induces the UPRER-immune response and avoidance behavior in C. elegans. Surprisingly, our findings indicate that supplementing HK-E. coli with carbohydrates such as D-Glc and D-GlcA does not promote animal development (Figure 3-figure supplement 2G), suggesting that carbohydrates are not essential for supporting animal growth on this food source. However, we did observe that carbohydrates play a critical role in inhibiting the UPRER-immune response induced by sugar deficiency in HK-E. coli.

-The authors claim that the induction of the innate immune pathway reporter irg-5::GFP is "abolished" in pmk-1(RNAi) animals, but Figure S2K seems to show a clear GFP signal when these animals are fed HK-OP50. Similarly, the claim that feeding WT animals HK-OP50 enriches phospho-PMK-1 levels (Fig 2E) is unconvincing - only one western blot is shown, with no quantification, and there is a smear in the critical first lane.

(1) The authors claim that the induction of the innate immune pathway reporter irg-5::GFP is "abolished" in pmk-1(RNAi) animals, but Figure S2K seems to show a clear GFP signal when these animals are fed HK-OP50.

We sincerely appreciate the reviewer's attention. To address this concern, we have replaced the images with higher resolution, larger ones in Figure 2-figure supplement 1-I. These updated images provide a clearer representation of the data, ensuring that all details are readily visible and enabling a more accurate interpretation of the results.

(2) Similarly, the claim that feeding WT animals HK-OP50 enriches phospho-PMK-1 levels (Fig 2E) is unconvincing - only one western blot is shown, with no quantification, and there is a smear in the critical first lane.

Thank you, following reviewer’s suggestion, we also repeated some of the western. We now replace the Figure 2E and quantified relative intensity of pPMK-1/tublin. We also provide the uncropped western blots images as source data ( “raw-data WB” file).

-The rationales for some of the paper's hypotheses could be improved. For example, the rationale for screening the E. coli mutant library is that some mutants, when heat-killed, may be missing a metabolite that induces the ER-UPR. A more straightforward hypothesis might be that some mutant E. coli strains aberrantly induce the ER-UPR when *not* heat-killed, because they are missing a metabolite that prevents stress pathway induction. This is not in itself a major concern, but it would be useful for the authors to provide a rationale for their hypothesis.

Thank you for the insightful suggestion. We acknowledge the importance of providing a clear rationale for our hypotheses in the paper. In response to this feedback, we have enhanced the discussion section to better elucidate the rationale behind our hypotheses.

One limitation of our study is the lack of explanation for why HK-E. coli activates UPRER and immunity. We hypothesized that when heat-killed, HK-E. coli may lack or contain altered levels of certain metabolites that either activate or inhibit UPRER and immunity, respectively. Additionally, we speculated that E. coli mutants killed by heat may lack metabolites that activate UPRER and immunity, or conversely, have increased levels of metabolites that inhibit these pathways.

Fortunately, our investigation led to the discovery of the E. coli mutant yfbR, which inhibits UPRER and immunity by increasing carbohydrates that aid in resisting these stress pathways. Moving forward, we intend to further explore the intricate relationship between HK-E. coli and UPRER-immunity. This will be a key focus of our future research efforts.

-The authors do not provide any explanation for some unexpected results from the E. coli screen. Earlier in the paper, the authors found that innate immune signaling is downstream of ER-UPR activation. However, of the 20 E. coli mutants that, when heat-killed, "did not induce... the UPR-ER reporter," 9 of them still activate the innate immune response. This seems at odds with the authors' simple model since it suggests that low-quality food can induce innate immune signaling independently of the ER-UPR. Further, only one of the 9 has an effect on behavior, even though failure to activate the innate immune pathway might be expected to lead to a behavioral defect in all of these.

Thank you for your understanding, and we apologize for any confusion caused by our earlier statement. To provide clarification, our study revealed that out of the 20 E. coli mutants examined, none activated the UPRER. Among these mutants, 9 did not induce immunity, and interestingly, one out of these 9 mutants demonstrated the ability to inhibit avoidance behavior.

This diversity in phenotypic outcomes can be attributed to the varied metabolites present in different E. coli mutants. To thoroughly evaluate the effects of these mutants, we conducted a comprehensive three-step screening process, utilizing UPRER marker, immunity marker, and avoidance behavior assays.

Through this rigorous approach, we identified the E. coli mutant, yfbR, which exhibited the desired inhibitory effects on UPRER, immunity, and avoidance behavior.

Subsequently, we conducted a metabolomics analysis of various food qualities (HK-K12, HK-yfbR, and Live-K12). Our findings revealed higher sugar levels in

HK-yfbR and Live-K12 compared to HK-K12 (Figure 3B, Figure 3-figure supplement 2A, and Table S1), indicating that sugar deficiency might trigger the UPRER, immunity responses, and subsequent avoidance behavior.

-In a number of places, the writing style can make the authors' arguments difficult to follow.

Thanks for the reviewer’s efforts. We changed all of these errors and polish the language of this paper.

-Some of the effect sizes observed by the authors are exceedingly small (e.g, the suppression of hsp-4::gfp induction by sugar supplementation in Figs 3C-E), raising some concern about the biological significance of the effect.

Thank you for your feedback. In response to your concern, we have included additional clarification in the manuscript.

We have added the following statement: “While sugar effectively inhibits the HK-E. coli-induced UPRER and immune response, it does not fully suppress it to the extent observed with live-E. coli (Figure 3C-F). This implies that additional nutrients present in live-E. coli might also contribute to the inhibition of UPRER and immune response.”

This addition helps to address the observation that some effect sizes appear small, providing context and suggesting potential factors that may influence the outcomes.

-In some cases, there is a discrepancy between the fluorescence images and their quantitation (e.g., Figure 3E, where the effect of glucose on GFP fluorescence seems much stronger in the image than in the graph).

Thank you for your valuable suggestion. In response, we have revised our image selection process to ensure impartiality. We now randomly select images to ensure they accurately represent the quantified data without bias. More details regarding this update can be found in Author response image 2.

More original picture corresponding to Figure 3E

Reviewer #3 (Public Review):

Summary:

Animals can evaluate food quality in many ways. In contrast to the rapid sensory evaluation with smell and taste, the mechanism of slow nutrient sensation and its impact on food choice is unexplored. The authors utilize C. elegans larvae and their bacterial food as an elegant model to tackle this question and reveal the detailed molecular mechanism to avoid nutrient-poor foods.

Strengths:

The strength of this study is that they identified the molecular identities of the critical players in bacterial food and C. elegans using unbiased approaches, namely metabolome analysis, E. coli mutant screening, and RNA sequencing. Furthermore, they strengthen their findings by thorough experiments combining multiple methods such as genetics, fluorescent reporter analysis, and Western blot.

Thank you for highlighting the strengths of our study.

Weaknesses:

The major caveat of this study is the reporter genes. The transcriptional reporters were used to monitor the UPRER and immune responses in the intestine of C. elegans.

However, their tissue-specific rescue experiments suggest that the genes in the UPRER and immune response function in the neurons. Thus, we should carefully interpret the results of the reporter genes.

Thank you for your insightful comment. We appreciate the opportunity to address your concerns regarding the interpretation of our reporter gene data.

Upon reevaluation, we observed strong induction of the UPRER reporter

(Phsp-4::GFP)8 and immunity reporter (Pirg-5::GFP)9 both in the intestine (Figure 1F-G) and in neurons (Figure 1-figure supplement 2A) in response to feeding unfavorable food (HK-E. coli). This suggests that both the UPRER and immune pathways may indeed respond to low-quality food (HK-E. coli) in multiple tissues of C. elegans. While we acknowledge that our tissue-specific rescue experiments suggest a role for these pathways in neurons, the intestinal fluorescence of Phsp-4::GFP or Pirg-5::GFP is easily observable and scorable. Therefore, we chose to focus our further analyses on the intestine for practical reasons.

Overall, this work provides convincing data to support their model. In the C. elegans field, the behaviors of larvae are not well studied compared to adults. This work will pose an interesting question about the difference between larvae and adults in nutrition sensing in C. elegans and provide a framework and candidate molecules to be studied in other organisms.

Recommendations for the authors:

Reviewer #1 (Recommendations For The Authors):

Major suggestions:

(1) My major overall comment is that the paper would be substantially strengthened by more mechanistic analysis. In particular, how does low vitamin C trigger UPR activation? This is an intriguing finding and it would be important to see it more fully explored.

Our study revealed that the vitamin C content in HK_-E. coli_ is comparable to that of live E. coli or HK-yfbR (Figure 4-figure supplement 1A), suggesting that the induction of unfolded protein response (UPR) in C. elegans by HK-E. coli is not attributed to low vitamin C levels, but rather to unknown factors.

Metabolomic analysis showed that the sugar levels, including lactose, D-(+)-sucrose, and D-(+)-glucose, were significantly decreased in HK-E. coli (Figure 3B, Table S1).

Furthermore, we found that supplementing D-(+)-glucose effectively inhibited UPRER (Figure 3E), immune response (Figure 3F, 3G, and Figure 3-figure supplement 2D), and avoidance behavior (Figure 3H) induced by HK-E. coli. Our findings suggest that the deficiency in sugars in HK-E. coli triggers a stress response and avoidance behavior in animals, which can be alleviated by D-(+)-glucose supplementation.

Notably, when E. coli was heat-killed, we observed that the sugar levels, including lactose and D-(+)-sucrose, were significantly higher in the heat-killed E. coli mutant yfbR (HK-yfbR) compared to HK-E. coli (Figure 3B). Moreover, we found that UPRER was reduced in animals feeding HK-yfbR (Figure 3-figure supplement 1B), indicating that higher sugar levels may inhibit the induction of UPRER by low-quality food.

The synthesis of vitamin C (VC) occurs through the glucuronate pathway, utilizing D-glucose as a precursor 1, 2 (Figure 4A). This led us to investigate whether the vitamin C biosynthesis pathway is involved in evaluating low-quality food by using D-glucose. In this study, we found that animals feeding live E. coli, which should produce more VC, exhibit higher glucose levels. However, our results show that animals maintain similar VC levels when fed ideal food (live E. coli) compared to low-quality food (HK-E. coli) (Figure 4B), suggesting that animals do not stimulate VC biosynthesis under favorable food conditions. In contrast, when animals are fed low-quality food (HK-OP50), we found that supplementing D-GlcA (Figure 4C) or E. coli-yfbR mutation (Figure 4B) in HK-E. coli can improve VC levels. Moreover, we found that VC or D-glucuronate (D-GlcA) supplementation inhibited HK-E. coli induced UPRER (Figure 4D). These data indicate that glucose boosts the animal's ability to adapt to unfavorable food environments by increasing VC levels, thereby inhibiting UPRER, but not in favorable food conditions.

In addition,we asked whether high level of VC inhibits other stress induced UPRER. Previous study shown that Tunicamycin6 and pathogenic bacteria-Pseudomonas aeruginosa-PA145 induce UPRER in C. elegans. We found that VC supplementation does not inhibit Tunicamycin or PA14-induced URPER (Author response image 3).

VC supplementation does not inhibit Tunicamycin or PA14-induced UPRER.

In addition, how is the activation of the PMK-1 pathway driven by/coordinated with UPR activation?

If the authors do not want to pursue these directions experimentally in this study, the discussion would be strengthened by considering these questions and identifying candidate regulatory mechanisms for further exploration.

In this study, we found that heat-killed E. coli (HK-E. coli), a low-sugar food, triggers cellular unfolded protein response (UPRER) and immune response. We also demonstrated that 1) the activation of UPRER by low-quality food depends on the IRE-1/XBP-1, 2) activation of immune response (PMK-1) is downstream of XBP-1 in responding to low-quality food.

how is the activation of the PMK-1 pathway driven by/coordinated with UPR activation?

In our discussion part, we added new data and discussion to answer reviewer’s question.

A previous study has shown that activating innate immunity (PMK-1 MAPK) leads to a reduction in translation 3. Our own previous research has also demonstrated that PMK-1 activation causes a shutdown of food digestion in animals4, likely to reduce protein translation and cellular metabolism. To investigate this further, we measured the translation level of animals fed with HK-E. coli and found that total translation ability is significantly reduced in these animals (Figure 5-figure supplement 1D). This finding suggests that activating innate immunity (PMK-1 MAPK) may serve as a mechanism to slow down translation progress, thereby alleviating the pressure on the unfolded protein response (UPR) and preventing excessive UPRER activation.

(2) Figure 2C: The data shows that xbp-1 mutants are significantly more likely to leave heat-killed E. coli. However, no other conditions are examined. Is this avoidance defect specific to heat-killed E. coli, or is it a more general effect of xbp-1 mutants - that is, are other conditions that evoke avoidance also affected by mutation of xbp-1? Is feeding behavior on regular E. coli altered in this background? The finding would be more relevant if the authors could clarify or provide more context for their claims here.

We then asked whether UPRER (IRE-1/XBP-1) - Innate immunity (PMK-1/p38 MAPK) axis is specific to evaluate low-quality food (HK-E. coli). We examined the avoidance behavior phenotype of wild-type and mutant L1 animals by placing them on various food conditions, including normal E. coli food, inedible food (Saprophytic staphylococci) and pathogenic food (Pseudomonas aeruginosa-PA14), for a 24-hour period. We found that N2, pmk-1, and xbp-1 mutant worms did not exhibit avoidance behavior when presented with normal food (OP50). However, both N2 and xbp-1 mutant worms were able to escape from inedible food, Saprophytic staphylococci, whereas pmk-1 mutant worms did not show this avoidance. Notably, xbp-1 mutant worms exhibited even more pronounced avoidance behavior when exposed to Pseudomonas aeruginosa, whereas pmk-1 mutant worms were more susceptible to infection by this pathogen (Figure 2-figure supplement 2C). These findings suggest that the UPR-Immunity pathway plays a crucial role in helping animals avoid low-quality food by triggering an avoidance response. In contrast, the Innate immunity pathway, which is mediated by PMK-1/p38 MAPK, appears to play a key role in evaluating unfavorable food sources, such as HK-E. coli, Saprophytic staphylococci, and Pseudomonas aeruginosa, and helping animals avoid these environments.

(3) Figure 3C-F: The magnitude of the changes between conditions shown in these panels is small. To what extent does this supplementation represent a full rescue? The findings would be strengthened if figures/images for the control condition (non-HK E. coli) were shown for comparison to allow the reader to assess the extent to which UPR/PMK-1 activation is rescued.

In response to a reviewer's suggestion, we included live-E. coli as a control in our study. Notably, our data revealed that the addition of lactose, D-(+)-sucrose, and D-(+)-glucose partially inhibited the HK-E. coli-induced unfolded protein response (UPRER) and immune response, suggesting that other nutrients present in live-E. coli may also play a role in inhibiting UPRER.

We added this in manuscript: “While sugar effectively inhibits the HK-E. coli-induced UPRER and immune response, it does not fully suppress it to the extent observed with live-E. coli (Figure 3C-F). This implies that additional nutrients present in live-E. coli might also contribute to the inhibition of UPRER and immune response.”

(4) Figure 5B-D: The magnitude of changes shown between conditions here again appear to be very small, even those labelled as statistically significant. It is important to ensure that the correct statistical tests have been used to assess the significance of these differences (see below).

All statistical analyses were performed in Graphpad prism 8.0. Two-tailed unpaired t test was used for statistical analysis of two groups of samples，one-way or two-way ANOVA was used for statistical analysis of more than two groups of samples.

(5) Methods: In the "Statistical analysis" section, the authors state that "All statistical analyses were performed using Student's t-test". However, this is not the appropriate test to use in experiments where multiple comparisons are made, which is true in several instances across the paper. In these cases, a more appropriate statistical test should be used.

All statistical analyses were performed in Graphpad prism 8.0. Two-tailed unpaired t test was used for statistical analysis of two groups of samples，one-way or two-way ANOVA was used for statistical analysis of more than two groups of samples.

Minor suggestions:

(1) Figure S2: RNAi is usually delivered in a different E. coli strain, HT115. Is this the case with the RNAi knockdowns in Figure S2, and given that diet can influence UPR activation, is it possible that this different diet could change the phenotypes observed?

This should be clarified by the authors.

In this study, all RNAi experiments involved bleaching adult animals under RNAi strain culture conditions to obtain L1 animals. Subsequently, L1 animals were transferred to HK-E. coli OP50 for phenotype analysis. In response to a reviewer's suggestion, we observed that L1 animals obtained from mothers fed E. coli strains OP50, HT115, or K12 exhibited similar UPR induction under HK-E. coli OP50 feeding conditions (Author response image 4). These findings suggest that variations in diet did not alter the UPR phenotypes.

L1 animals obtained from mothers fed E. coli strains OP50, HT115, or K12 exhibited similar UPR induction under HK-E. coli OP50 feeding conditions

Reviewer #2 (Recommendations For The Authors):

Line 182: "irg-5::GFP" should be "hsp-4::gfp".

Thanks for the reviewer’s efforts. We have changed this error.

Reviewer #3 (Recommendations For The Authors):

Major comments:

(1) The reporter genes of UPRER and immune response were analyzed in the intestine throughout the study. On the other hand, their rescue experiments suggest that these pathways function in the neurons. They should provide the fluorescence data in the neurons at least for Figures 1F and 1G to confirm that the intestinal response matches the neuronal response and mention that further analyses were done in the intestine for easy scoring.

Consistent with the results of the RNA sequencing (RNA-seq) analysis, the UPRER reporter (Phsp-4::GFP)8 and immunity reporter (Pirg-5::GFP)9 were strongly induced in intestinal (Figure 1F-G) and neurons (Figure 1-figure supplement 2A) by feeding unfavorable food (HK-E. coli), suggesting that UPRER and immune pathways may respond to low-quality food (HK-E. coli). As intestinal fluorescence (Phsp-4::GFP or Pirg-5::GFP) is easy observation and scoring, the further analyses were done in the intestine.

(2) I have concerns about the interpretation of the p-PMK-1 data. Although the authors described that "p-PMK-1 is prominently increased" in the text (Line 150), it is unclear on the data (Figure 2E). Similarly, the authors' statement "p-PMK-1 is decreased in animals with D-GlcA (F).." was not fully supported by the data in Figure 4F. The experiment should be repeated and quantified. Moreover, pPMK-1 showed single bands in Figure 2E, but double bands in Figure 3G, 4F, and 4G. The authors should explain why that is the case and which band we should look at for Figures 3G, 4F, and 4G.

As reviewer’s suggestion, we also repeated some of the western. We found that after longer expose, there are two bands for pPMK-1 (Figure 2E, new data; and “raw-data WB” file). The VHP-1 phosphatase is known to inhibit PMK-13. In our previous study, we found that worms treated with vhp-1(RNAi), which hyperactivates p-PMK-1 (lower band) 4. In contrast, the two bands are disappeared in pmk-1 mutant (Author response image 5). Thus, the lower band indicates the pPMK-1. We now replace the Figure 2E and quantified relative intensity of pPMK-1/tublin. We also provide the uncropped western blots images as source data ( “raw-data WB” file).

In our previous study, we found that worms treated with vhp-1(RNAi), which hyperactivates p-PMK-1 (lower band) 4.

In contrast, the two bands are disappeared in pmk-1 mutant. These pictures are extracted from our previous study4.

(3) Heat-killed E. coli (HK-E. coli) is low-quality because the lack of sugar cannot support the growth of C. elegans larvae (Qi and Han, Cell, 2018). Thus, animals do not show the UPRER-immune response and avoidance when HK-E. coli is supplemented with sugars such as glucose (Line 225-227). If these sugars are the key, C. elegans larvae should be able to grow better with HK-E. coli supplemented with glucose. Authors should address this possibility.

Previous studies have shown that heat-killed E. coli (HK-E. coli) is a low-quality food source that cannot support the growth of C. elegans larvae7. Here, we found that sugar deficiency in HK-E. coli induces the UPRER-immune response and avoidance behavior in C. elegans. Given this, we investigated whether sugar supplementation could promote animal growth when fed HK-E. coli. To our surprise, supplementing HK-E. coli with carbohydrates (D-Glc, D-GlcA) did not support animal development (Figure 3-figure supplement 2G), suggesting that carbohydrates are not essential for supporting animal growth on this food source. However, we did find that carbohydrates are critical for inhibiting the UPRER-immune response induced by sugar deficiency in HK-E. coli.

(4) Line 884: Instead of the Student's t-test, the ANOVA should be used for multiple comparisons.

All statistical analyses were performed in Graphpad prism 8.0. Two-tailed unpaired t test was used for statistical analysis of two groups of samples，one-way or two-way ANOVA was used for statistical analysis of more than two groups of samples.

(5) Although the results are interesting and convincing, the manuscript needs some careful editing and proofreading. As far as I could catch, there are more than 100 errors and typos, as I summarized in minor comments. I recommend the authors proofread thoroughly to make this work easier to read.

Thanks for the reviewer’s efforts. We changed all of these errors and polish the language of this paper.

Minor comments:

(1) Line 30: nature -> natural

(2) Line 86: elegnas -> elegans

(3) Line 93: the17h -> the 17h

(4) Line 97: response -> respond

(5) Line106: responded -> respond

(6) Lien 107-109: Add references for the three reporters

(7) Line 114: immune -> immune pathway

(8) Line 118: immune depended -> immune-dependent

(9) Line 128, 594, 596: deferentially -> differentially

(10) Line 131: Explain what IRE-1-mediated splicing of xbp-1 with references

(11) Line 170: XPB-1 -> XBP-1

(12) Line 179: URP -> UPR

(13) Line 181: hsp-4::GFP -> Phsp-4::GFP

(14) Line 183: Italicize E. coli; mutant -> mutants

(15) Line 184: irg-5::GFP -> Pirg-5::GFP (2 places)

(16) Line 197, 203, 206, 207: Lactose -> lactose

(17) Line 206, 209, 217, 225, 228, 232, 237, 262, 442, 445, 604, 739: Glucose -> glucose

(18) Line 218: Sugars deficiency -> sugar deficiency

(19) Line 229: found contribute to -> found to contribute to

(20) Line 235, 537, 539, 587, 599, 642, 855: Italicize E. coli

(21) Line 236: same -> the same

(22) Line 239: I recommend adding "in C. elegans". This study uses both E. coli and C.

elegans genetics. Sometimes, it is confusing which organism was mentioned. It should be applied where it is necessary.

(23) Line 240: additional -> addition

(24) Line 339, 642: Italicize kgb-1

(25) Line 390: Italicize Pseudomonas aeruginosa, Bacillus thuringiensis,

Staphylococcus aureus, and Serratia marcescens

(26) Line 394: wiht -> with

(27) Line 400, 550: Change ER to superscript; Italicize ire-1, xbp-1, and pmk-1

(28) Line 415: xpb-1 -> xbp-1

(29) Line 460, 525, 531, 532, 617, 655: Italicize yfbR

(30) Line 457, 468, 472, 475, 482, 497, 513, 624, 629, 633, 733. 758: Vitamin -> vitamin

(31) Line 459: Make it clear what is the relationship between vitamin C and TAA

(32) Line 527: Do not italicize mutant

(33) Line 538: Phsp-6:GFP -> Phsp-6::GFP (to match other descriptions)

(34) Line 540: Phsp-4:GFP -> Phsp-4::GFP (to match other descriptions)

(35) Line 540: Italicize hsp-4

(36) Line 543: Pirg-5:GFP -> Pirg-5::GFP (to match other descriptions) and italicize irg-5

(37) Line 550, 881: Innate -> innate

(38) Line 557, 560, 564, 838: Do not italicize HK

(39) Line 561: Remove the extra space before "three"

(40) Line 575, 577: Reporter -> reporter

(41) Line 575, 607: Italicize Phsp-4::GFP

(42) Line 577: immunity -> Immunity; Italicize Pirg-5::GFP

(43) Line 585, 653: keio -> Keio

(44) Line 586: hsp-4::GFP -> Phsp-4::GFP

(45) Line 586, 589 (2 places): irg-5::GFP -> Pirg-5::GFP

(46) Line 597: Remove "all"

(47) Line 600: Trehalose -> trehalose

(48) Line 609: Italicize Pirg-5::GFP

(49) Line 615: critically -> critical

(50) Line 636: Remove "+"

(51) Line 656 (2 places), 682: Do not italicize OP50

(52) Line 664: Lead -> lead

(53) Line 681: Describe the composition of NGM or show the reference. Since this paper examines nutrition, the composition of the medium is crucial.

(54) Line 686-706: Italicize all allele names. Be consistent with how to write the promoter to avoid confusion (e.g., ttx-3p -> Pttx-3). Be consistent with how to describe the transgene (e.g., Phsp-4::GFP(zcIs4) -> zcIs4[Phsp-4::GFP])

(55) Line 710: Describe the composition of LB or show the reference. Since this paper examines nutrition, the composition of the medium is crucial.

(56) Line 709, 856 (2 places), 858: Do not italicize K12 to make it consistent

(57) Line 719: Podr-1p:RFP -> Podr-1::RFP

(58) Line 722, 724: Italicize ges-1 and xbp-1

(59) Line 723: Pges-1:xbp-1::GFP -> Pges-1::xbp-1::GFP

(60) Line 735: Glucuronic -> glucuronic

(61) Line 748: I believe it is 5 mm instead of 0.5 mm

(62) Line 750: The equation should be (5 mm)2/(17.5 mm)2

(63) Line 759: Remove the period after "pattern".

(64) Line 766: Describe how they were synchronized

(65) Line 774: Italicize Psysm-1p::GFP

(66) Line 785: Insert a space before "until"

(67) Line 787: the mutant -> mutant

(68) Line 789, 792, 793, 795 (2 places): GPF -> GFP

(69) Line 791: next -> Next; an -> a

(70) Line 799: Remove a space before "MRC".

(71) Line 804: I do not understand what "until adulthood" means in this context;

Remove a space before "by". (I recommend searching double space and correcting it.)

(72) Line 853: Metabolome -> metabolome

(73) Line 893-1082: Species and gene names should be italicized in Reference

(74) Figures 1F, 1G, S2F, S2G: The panels' order should match the bar graphs' order. The apparent difference in the representative data does not match the marginal difference in the bar graph in Fig. 1G. The authors should double-check the results.

(75) Figure 1F, 2A, 2B, 3C, 3D, 3E, 4D, 4I, S1J, S2A, S2B, S2I, S3B, S3F, S3H: hsp-4::GFP -> Phsp-4::GFP

(76) Figure 1G, 2D, 3F, 4E, 4J, S1K, S2H, S3C, S3I: irg-5::GFP -> Pirg-5::GFP

(77) Figure 6: Liquids -> Lipids; Italicize ire-1, xbp-1, pmk-1

(78) Figure S1I: hsp-6::GFP -> Phsp-6::GFP

(79) In the legend for Figure S1 after Figure S1, (A), (B)... were duplicated. It is OK in the corresponding main text (Line 530)

(80) Figure S2F, S3G, S4C, S4D: sysm-1::GFP -> Psysm-1::GFP

(81) Figure S2G: irg-1::GFP -> Pirg-1::GFP

(82) Figure S3H and S3I: Describe which ones are Glu + conditions

References:

(1) Patananan AN, Budenholzer LM, Pedraza ME, Torres ER, Adler LN, Clarke SG. The invertebrate Caenorhabditis elegans biosynthesizes ascorbate. Arch Biochem Biophys 569, 32-44 (2015).

(2) Yabuta Y_, et al._ L-Ascorbate Biosynthesis Involves Carbon Skeleton Rearrangement in the Nematode Caenorhabditis elegans. Metabolites 10, (2020).

(3) Weaver BP, Weaver YM, Omi S, Yuan W, Ewbank JJ, Han M. Non-Canonical Caspase Activity Antagonizes p38 MAPK Stress-Priming Function to Support Development. Dev Cell 53, 358-369 e356 (2020).

(4) Geng S_, et al._ Gut commensal E. coli outer membrane proteins activate the host food digestive system through neural-immune communication. Cell Host Microbe 30, 1401-1416 e1408 (2022).

(5) Richardson CE, Kooistra T, Kim DH. An essential role for XBP-1 in host protection against immune activation in C. elegans. Nature 463, 1092-1095 (2010).

(6) Harding HP_, et al._ An Integrated Stress Response Regulates Amino Acid Metabolism and Resistance to Oxidative Stress. Molecular Cell 11, 619-633 (2003).

(7) Qi B, Kniazeva M, Han M. A vitamin-B2-sensing mechanism that regulates gut protease activity to impact animal’s food behavior and growth. eLife 6, e26243 (2017).

(8) Calfon M_, et al._ IRE1 couples endoplasmic reticulum load to secretory capacity by processing the XBP-1 mRNA. Nature 415, 92-96 (2002).

(9) Bolz DD, Tenor JL, Aballay A. A Conserved PMK-1/p38 MAPK Is Required in Caenorhabditis elegans Tissue-specific Immune Response to Yersinia pestis Infection*. The Journal of Biological Chemistry 285, 10832 - 10840 (2010).

Version 2

Abstract

Abstract

To survive in challenging environments, animals must develop a system to assess food quality and adjust their feeding behavior accordingly. However, the mechanisms that regulate this chronic physiological food evaluation system, which monitors specific nutrients from ingested food and influences food-response behavior, are still not fully understood. Here, we established a low-quality food evaluation assay system and found that heat-killed E. coli (HK-E. coli), a low sugar food, triggers cellular UPRER and immune response. This encourages animals to avoid low-quality food. The physiological system for evaluating low-quality food depends on the UPRER (IRE-1/XBP-1) – Innate immunity (PMK-1/p38 MAPK) axis, particularly its neuronal function, which subsequently regulates feeding behaviors. Moreover, animals can adapt to a low-quality food environment through sugar supplementation, which inhibits the UPRER –PMK-1 regulated stress response by increasing vitamin C biosynthesis. This study reveals the role of the cellular stress response pathway as physiological food evaluation system for assessing nutritional deficiencies in food, thereby enhancing survival in natural environments.

Introduction

Food is essential for the survival, growth and fitness of all animals. To adapt to fluctuating environments with a wide range of food sources, animals have developed a food evaluation system. This system enables them to identify nutrient-rich food and avoids low-quality or toxic food, thereby maximizing their survival prospects 1, 2, 3. Various sensory neuron evaluation systems in animals has evolved to evaluate food quality through vision 4, 5, olfactory 3, 6, 7, 8, 9, 10, 11,, 12 and gustatory senses 13, 14, 15. Besides these sensory systems that facilitate quick feeding decisions, animals may also initiate cellular stress response programs to detect nutrition/toxin and trigger food response behaviors 16, 17. This could be one of physiological food quality evaluation systems that monitor the nutritional status of consumed food. However, the signaling events in cellular stress responses involved in evaluating of specific nutrients and the mechanisms that connect these signaling activities to food behaviors are largely unexplored. More specifically, while cellular stress response through UPRER 18 and PMK-1/p38 MAPK 19 dependent immunity in response to pathogens have been extensively studied, the functions of these cellular stress response in sensing and evaluating specific nutrients from food remain unclear.

Vitamin C is an essential micronutrient that cannot be synthesized by humans due to the loss of a key enzyme in the biosynthetic pathway 20. Animals obtain vitamin C from their diet and possibly also from gut microbes 21. Vitamin C is an important physiological antioxidant and a cofactor for a family of biosynthetic and gene regulatory monooxygenase and dioxygenase enzymes. It is also required for the biosynthesis of collagen, L-carnitine, and certain neurotransmitters 22, 23. Vitamin C has been associated with various human diseases including scurvy, immune defect and cardiovascular disease 20. Therefore, in animals, having robust food evaluation systems to detect vitamin C levels could significantly impact their survival in the wild. However, the potential involvement of the cellular stress response pathway in this food evaluation system for sensing and assessing vitamin C remains largely unexplored.

In this study, using the low-quality food evaluation assay system we established 24, we elucidated the mechanism by which the cellular stress response pathway operates as a physiological food evaluation system. This pathway assesses the deficiency of D-glucose in food and the subsequent vitamin C content in animals through the unfolded protein response (UPRER) – innate immunity (PMK-1/p38 MAPK) axis. This mechanism promotes animals to leave low-quality food and is critical for their survival in nature environments.

Results

Low-quality food induces stress response in animals

Our previous studies have shown that Heat-killed E. coli (HK-E. coli), which lacks certain molecules, is considered a low-quality food that is unable to support animal growth 24, 25. Moreover, through metabolic-seq analysis, we identified significant changes in a large numbers of derivatives (Figure 1 – figure supplement 1A, Table S1), including lipids and their derivatives (Figure 1-figure supplement 1B, Table S1), amino acid and their metabolites (Figure 1-figure supplement 1C, Table S1), as well as coenzymes and vitamins (Figure 1-figure supplement 1D, Table S1). Interestingly, we observed a significant decrease in carboxylic acids and their derivatives (Figure 1-figure supplement 1E, Table S1) in E. coli after being heat-killed (Figure 1-figure supplement 1F, Table S1). This suggests that HK-E. coli is nutritionally deficient for C. elegans when compared to normal E. coli food.

Next, we conducted two behavior assays to facilitate the analysis of the food evaluation process in animals by seeding L1 animals in assay plates (Figure 1A and 1B). In the avoidance assay, wild-type animals avoided the HK-E. coli OP50 (HK-OP50) food (Figure 1A). Interestingly, in the food choice assay, animals initially showed no preference between the two types of food (1-2h), but eventually exhibited a preference for high-quality food (Live E. coli) up until the 17h mark (Figure 1B, Figure 1-figure supplement 1G). This suggests that worms depart from the HK-E. coli after recognizing it as low-quality food source through ingestion.

The stress response is induced in animals fed low-quality food, HK-E. coli.

(A) Schematic drawing and quantitative data of the food aversion assay. Circles indicate the food spot for live (yellow) and HK-OP50 (red) bacteria, respectively. The animals were scored 16-17 hours after L1 worms were placed on the food spot. Data are represented as mean ±SD from three independent experiments, 79-129 animals/assay.

(B) Schematic method and quantitative data of the food selection assay. Live (yellow), heat-killed (red) E. coli and LB as the buffer for E. coli were placed on indicated position. Synchronized L1 worms were place in the center spot. The selection index was calculated at the indicated time. Data are represented as mean ±SD from eight independent experiments, 123-792 animals/assay.

(C) GO enrichment analysis of up-regulated genes in animals fed with HK-E. coli vs live E. coli.

(D) Venn diagram showing numbers of UPRER target genes and up-regulated genes in animals fed HK-E. coli, and their overlap.

(E) Venn diagram showing numbers of induction genes by four pathogenic bacteria and HK-E. coli induced genes, and their overlap. The gene expression data was extracted from published data of animals’ infection with Pseudomonas aeruginosa (PA) 26, Bacillus thuringiensis (Bthu) 27, Staphylococcus aureus (Saur) 27, and Serratia marcescens (Smar) 27.

(F-G) GFP fluorescence images and bar graph showing that Phsp-4::GFP (F) and Pirg-5::GFP (G) were induced in animals fed with HK-E. coli. Blue numbers are the number of worms scored from at least three independent experiments. Data are represented as mean ± SD.

For all panels, Scale bar shows on indicated figures, 50 μm. * p<0.05, ** p<0.01, *** p < 0.001, **** p<0.0001, ns: no significant difference. Precise P values are provided in Raw Data.

In order to investigate the pathways in animals that respond to HK-E. coli, we performed transcriptomics analysis on worms that were cultured with both HK-E. coli and Live E. coli. Gene-expression profiling revealed that stress response genes, including those related to biotic stimulus, immune response and response to stress, are up-regulated in animals fed with HK-E. coli OP50 (HK-OP50) (Figure 1C, Table S2). Among these up-regulated genes, we identified 11 out of 62 of UPRER target genes (Figure 1D, Figure 1-figure supplement 1H and Table S2). Additionally, about 50%-80% of up-regulated genes overlap with genes responding pathogenic bacteria 26, 27 (Figure 1E, Table S2). Consistent with the results of the RNA sequencing (RNA-seq) analysis, the UPRER reporter (Phsp-4::GFP)28 and immunity reporter (Pirg-5::GFP)29 were strongly induced in intestine (Figure 1F-G) and neurons (Figure 1-figure supplement 2A) by feeding unfavorable food (HK-E. coli OP50), suggesting that UPRER and immune pathways may respond to low-quality food (HK-E. coli OP50). As intestinal fluorescence (Phsp-4::GFP or Pirg-5::GFP) is easy observation and scoring, the further analyses were done in the intestine.

Moreover, UPRMt reporter (Phsp-6::GFP) 30 was weakly induced under HK-E. coli feeding condition (Figure 1-figure supplement 2B), and starved worm did not induce UPRER and immunity (Figure 1-figure supplement 2C-D).

Together, these findings suggest that low-quality food (HK-E. coli OP50) triggers a stress response pathway in animals, including UPRER and innate immune pathway. This implies that animals may assess the quality of food through UPRER and innate immune pathway.

Animals evaluate food quality through UPRER-immune-dependent physiological food quality evaluation system

To determine whether the UPRER and innate immune pathways play a role in evaluating low-quality food, we first examined whether the activation of the UPRER by HK-E. coli was dependent on the known signaling components of the UPRER branches, including IRE/XBP-1, PERK/ATF-4 and ATF-6 31, 32. We observed no difference in Phsp-4::GFP induction with atf-4 (Figure 2-figure supplement 1A) or atf-6 (Figure 2-figure supplement 1B) RNAi-mediated knockdown in animals fed with HK-E. coli. However, knockdown of ire-1/xbp-1 or mutation of xbp-1 reduced GFP fluorescence (Figure 2A, 2B). Among the 11 differentially expressed UPRER target genes in animals fed with HK-E. coli from RNA-seq (Figure 1D), 64% of the genes are IRE-mediated genes (Figure 1 – figure supplement 1H, Table S2). The mRNA level of IRE-1-mediated splicing of xbp-128 is also induced in animals fed with HK-E. coli OP50 (Figure 2-figure supplement 1C). However, UPRER is not affect in animals feeding live-E. coli by RNAi of ire-1, xpb-1, atf-4 and atf-6 (Figure 2-figure supplement 1D). These data suggest that activation of the UPRER by low-quality food (HK-E. coli) depends on the IRE-1/XBP-1.

Animals evaluate food quality through UPRER (ire-1/xbp-1) – Innate immunity (pmk-1 MAPK) axis.

(A-B) GFP fluorescence images and bar graph showing that HK-E. coli induced Phsp-4::GFP was decreased in animals with ire-1 or xbp-1 RNAi treatment (A) or xbp-1 mutation (B). Blue numbers are the number of worms scored from at least three independent experiments. Data are represented as mean ± SD.

(C) Food aversion assay showing that xbp-1 mutation eliminated the discrimination against HK-E. coli. However, this effect is rescued by expressing xbp-1 in neurons rather than intestine. Data are represented as mean ±SD from five independent experiments, 156-763 animals/assay.

(D) GFP fluorescence images and bar graph showing that HK-E. coli induced Pirg-5::GFP was decreased in animals with pmk-1 RNAi treatment. Blue numbers are the number of worms scored from at least three independent experiments. Data are represented as mean ± SD.

(E) Western blot images showing the level of p-PMK-1 in L1 animals (Wild-type N2 and xbp-1 mutant) fed with OP50 or HK-OP50 for 4 h. The level of p-PMK-1 is induced in animals fed HK-OP50.

(F) Food aversion assay showing that pmk-1 mutation eliminated the discrimination against HK-E. coli. However, this effect is rescued by expressing pmk-1 in neurons rather than intestine. Data are represented as mean ±SD from five independent experiments, 168-492 animals/assay.

(G) Food aversion assay in wild-type, xbp-1, pmk-1 and double mutant. Data are represented as mean ±SD from five independent experiments, 259-490 animals/assay.

For all panels, Scale bar shows on indicated figures, 50 μm. * p<0.05, ** p<0.01, *** p < 0.001, **** p<0.0001, ns: no significant difference. Precise P values are provided in Raw Data.

To further analyze whether XBP-1-dependent UPRER activation is critical for animals to leave low-quality food, we tested food avoidance behavior using xbp-1 mutant. The results show that xbp-1 mutants had a significantly decreased likelihood of leaving of HK-E. coli, which was rescued by expressing xbp-1 in neuron rather than intestine (Figure 2C). This indicates that XBP-1-dependent UPRER activation in neuron is critical for animals to specific evaluate low-quality food (HK-E. coli).

We then investigated which innate immune pathway is involved in evaluating low-quality food. First, we analyzed HK-E. coli induced genes from RNA-seq. Among these up-regulated genes, 82 out of 409 of PMK-1-dependent genes 33 were identified (Figure 2-figure supplement 1E, Table S2). Second, we confirmed the induction of several well-known PMK-1 target genes in RNA-seq data34 (Figure 2-figure supplement 1F) and reporter analysis (Figure 2-figure supplement 1G-H). Moreover, the induction of Pirg-5::GFP was abolished in pmk-1 knockdown animals fed with HK-E. coli (Figure 2D). Third, we found that the phosphorylated PMK-1 (p-PMK-1) level was prominently increased in wild-type N2 animals fed HK-E. coli compared to feeding E. coli OP50 (Figure 2E). Finally, pmk-1 mutants had a decreased likelihood of leaving of HK-E. coli, which was rescued by expressing pmk-1 in neurons rather than intestine (Figure 2F). Moreover, Pirg-5::GFP is not affect in animals feeding live-E. coli by RNAi of pmk-1 (Figure 2-figure supplement 1I).These data suggest that PMK-1 regulated immune pathway evaluates low-quality food, especially the neuronal PMK-1 has a critical function for food quality response.

XBP-1 and PMK-1 are in the same pathway for evaluating food quality

Next, we explored the connection between UPRER (IRE-1/XBP-1) and innate immunity (PMK-1 p38 MAPK) in food quality evaluation. We found that Pirg-5::GFP induction (Figure 2-figure supplement 1I, 2A) and PMK-1 activation (Figure 2E) were decreased in animals with xbp-1 mutation or knockdown when fed with HK-E. coli, suggesting that XBP-1 could regulate PMK-1 under this condition. Additionally, Phsp-4::GFP induction under HK-E. coli was not affected in animals with pmk-1 RNAi (Figure 2-figure supplement 1D, 2B), indicating that XBP-1-dependent UPRER activation is not regulated by PMK-1. Finally, we constructed a double mutant of xbp-1 and pmk-1 and found that the food avoidance phenotype of the double mutant was similar to the pmk-1 mutant (Figure 2G), indicating that PMK-1 is downstream of XBP-1 in responding to low-quality food.

We then asked whether UPRER (IRE-1/XBP-1) – Innate immunity (PMK-1/p38 MAPK) axis is specific to evaluate low-quality food (HK-E. coli). We performed behavior assay in N2, pmk-1 and xbp-1 mutant animals by feeding normal E. coli food, inedible food (Saprophytic staphylococci) 35 and pathogenic food (Pseudomonas aeruginosa-PA14) 18. We found that N2, pmk-1, and xbp-1 mutant worms did not exhibit avoidance behavior when presented with normal food (OP50). However, both N2 and xbp-1 mutant worms were able to escape from inedible food (N2 was predominantly found on the border areas of the bacterial lawn and xbp-1 mutant worms on border and in), Saprophytic staphylococci, whereas pmk-1 mutant worms did not exhibit this avoidance behavior. Notably, N2 and xbp-1 mutant worms exhibited even more pronounced avoidance behavior when exposed to Pseudomonas aeruginosa, whereas pmk-1 mutant worms were more susceptible to infection by this pathogen (Figure 2-figure supplement 2C). These findings suggest that the UPR-Immunity pathway plays a crucial role in helping animals avoid low-quality food (HK-E. coli) by triggering an avoidance response. In contrast, the Innate immunity pathway, which is mediated by PMK-1/p38 MAPK, appears to play a key role in evaluating unfavorable food sources, such as HK-E. coli, Saprophytic staphylococci, and Pseudomonas aeruginosa, and helping animals avoid these environments.

Sugar deficiency in HK-E. coli food induces stress response and avoidance behavior in animals

We then investigated which nutrients/metabolites are sensed by animals through the XBP-1-PMK-1 axis for food quality evaluation. First, we hypothesized that the nutrition status is improved in E. coli mutant (HK-treatment), which could inhibit UPRER and immune response in animals. We established a system for screening the E. coli mutant Keio library (Figure 3 – figure supplement 1A), and identified 20 E. coli mutants that did not induce Phsp-4::GFP through the UPRER reporter (Phsp-4::GFP) after three rounds of screening (Table S3). From these 20 E. coli mutants, we identified 9 E. coli mutants that did not induce Pirg-5::GFP through the immunity reporter (Pirg-5::GFP) screening (Figure 3-figure supplement 1B-C, Table S3). Animals fed HK-yfbR, which catalyzes carbohydrate derivative metabolic process36, had a decreased ability to leave food (Figure 3A, Table S3), indicating that HK-yfbR may be a higher quality food for animals compared to HK-K12.

HK-E. coli is low sugar food, which induce stress response and avoidance behavior in animals.

(A) Food aversion assay showing that wild-type animals eliminated the discrimination against HK-E. coli when yfbR is mutated in E. coli. Data are represented as mean ±SD four independent experiments, 251-490 animals/assay.

(B) Heat map showing the thirteen differential metabolites from HK-K12, HK-yfbR, and K12 in 4 independent experiments. Color indicates the relative level of each metabolite.

(C-D) The bar graph showing that HK-E. coli induced Phsp-4::GFP was decreased in animals with lactose (C) or D-(+)-sucrose (D) supplementation. Blue numbers are the number of worms scored from at least three independent experiments. Data are represented as mean ± SD.

(E-F) GFP fluorescence images and bar graph showing that HK-E. coli induced Phsp-4::GFP (E) and Pirg-5::GFP (F) were decreased in animals with D-(+)-glucose (D-Glc) supplementation. Blue numbers are the number of worms scored from at least three independent experiments. Data are represented as mean ± SD.

(G) Western blot images showing the level of p-PMK-1 in L1 animals fed HK-E. coli with or without D-(+)-glucose (D-Glc) supplementation for 4 h. The level of p-PMK-1 is decreased in animals fed HK-OP50+D-Glc.

(H) Food aversion assay showing that wild-type animals eliminated the discrimination against HK-E. coli with D-Glc supplementation. Data are represented as mean ±SD six independent experiments, 190-492 animals/assay.

For all panels, Scale bar shows on indicated figures, 50 μm. * p<0.05, ** p<0.01, *** p < 0.001, **** p<0.0001, ns: no significant difference. Precise P values are provided in Raw Data.

Secondly, we performed a metabolomics analysis of different quality food (HK-K12, HK-yfbR and Live-K12). We found that the level of 13 metabolites were similar between HK-yfbR and Live-K12, but significantly changed in HK-K12 (Figure 3B, Figure 3-figure supplement 2A, and Table S1). We also found that genes involved in glycolysis/gluconeogenesis were up-regulated in animals fed with HK-E. coli (Figure 3-figure supplement 2B), suggesting that glycolysis/gluconeogenesis metabolism is disordered in animals fed with HK-E. coli, which may result from changes in sugar/carbohydrate intake. The carbohydrates (D-trehalose, lactose, and D-(+)-sucrose) were also decreased in HK-E. coli (Figure 3B), suggesting that carbohydrate deficiency may induce stress response and avoidance behavior in animals feeding HK-E. coli.

Thirdly, to determine which carbohydrate inhibits stress response in animals, we supplemented each metabolite to HK-E. coli and found that only Lactose, and D-(+)-sucrose inhibited HK-E. coli induced UPRER (Figure 3C and 3D, Figure 3-figure supplement 2C). Moreover, we found from our metabolomic data that the sugar level, including lactose, and D-(+)-sucrose, and D-(+)-glucose, was also decreased in HK-E. coli (Figure 3B, Table S1). Since lactose and D-(+)-sucrose are hydrolyzed to produce glucose37, 38, we wondered whether glucose also inhibits the stress response in animals. We found that D-(+)-glucose supplementation also inhibited HK-E. coli induced UPRER (Figure 3E), immune response (Figure 3F, 3G and Figure 3-figure supplement 2D) and avoidance (Figure 3H). Moreover, sugar supplementation did not affect UPRER and immunity in normal food (OP50) or starved condition (NGM) (Figure 3 – figure supplement 2E-F). While sugar effectively inhibits the HK-E. coli-induced UPRER and immune response, it does not fully suppress it to the extent observed with live-E. coli (Figure 3C-F). This implies that additional nutrients present in live-E. coli might also contribute to the inhibition of UPRER and immune response.

Previous studies have shown that heat-killed E. coli (HK-E. coli) is a low-quality food source that cannot support the growth of C. elegans larvae24, 25, whereas supplementation with vitamin B2 (VB2) can restore animal growth24. Here, we found that sugar deficiency in HK-E. coli induces the UPRER-immune response and avoidance behavior in C. elegans. Given this, we investigated whether sugar supplementation could promote animal growth when fed HK-E. coli. To our surprise, supplementing HK-E. coli with carbohydrates (D-Glc, D-GlcA) did not support animal development (Figure 3-figure supplement 2G), suggesting that carbohydrates are not essential for supporting animal growth on this food source. However, we did find that carbohydrates are critical for inhibiting the UPRER-immune response induced by sugar deficiency in HK-E. coli.

Together, these findings suggest that HK-E. coli induces a stress response and avoidance behavior in animals, which can be inhibited by D-(+)-glucose supplementation. This implies that animals may evaluate the sugar deficiency from HK-E. coli through the activation of UPRER and immune responses.

Animals could overcome a low-quality food environment by sugar supplementation through vitamin C biosynthesis

We discovered that D-(+)-glucose supplementation inhibited HK-E. coli induced UPRER (Figure 3E), immune response (Figure 3F, 3G and Figure 3-figure supplement 2D) and avoidance (Figure 3H). Simultaneously, vitamin C (VC), which is synthesized by glucuronate pathway using D-glucose 39, 40 (Figure 4A), was found to contribute to neuroprotective41, 42, immune defense43, 44, and inhibits inflammatory and ER stress45, 46. This led us to question whether the vitamin C biosynthesis pathway is involved in evaluating low-quality food by using D-glucose.

Vitamin C biosynthesis pathway is critically involved in evaluating sugar in the food

(A) Cartoon illustration of a simplified, Vitamin C biosynthesis pathway in C. elegans. The relevant coding genes of enzymes was labeled with red.

(B) The level of total L-ascorbic acid (TAA), also known as vitamin C, in animals fed with Live-K12, HK-K12, or HK-yfbR. Data are represented as mean ±SD from six independent experiments.

(C) The level of total L-ascorbic acid (TAA) in animals (control or knockdown of Vitamin C biosynthesis genes) fed with HK-E. coli with or without D-glucuronate (D-GlcA) supplementation. Data are represented as mean ±SD from eight independent experiments.

(D-E) GFP fluorescence images and bar graph showing that HK-E. coli induced Phsp-4::GFP (D) and Pirg-5::GFP (E) were decreased in animals with D-GlcA or Vitamin C supplementation. Blue numbers are the number of worms scored from at least three independent experiments. Data are represented as mean ± SD.

(F-G) Western blot images showing the level of p-PMK-1 in L1 animals fed with HK-E. coli with D-GlcA or Vitamin C supplementation for 4 h. The level of p-PMK-1 is decreased in animals with D-GlcA (F) or Vitamin C (G) supplementation.

(H) Food aversion assay showing that wild-type animals eliminated the discrimination against HK-E. coli with D-GlcA or Vitamin C supplementation. Data are represented as mean ±SD from four independent experiments, 153-292 animals/assay.

(I-K) The bar graph showing that suppression of HK-E. coli induced Phsp-4::GFP (I), Pirg-5::GFP (J) and food avoidance (K) by D-GlcA supplementation was abolished in animals with RNAi of VC biosynthesis genes, which was not affect by Vitamin C supplementation. Blue numbers are the number of worms scored from at least three independent experiments and Data are represented as mean ±SD(I-J). Data are represented as mean ±SD from five independent experiments, 252-537 animals/assay (K).

For all panels, Scale bar shows on indicated figures, 50 μm. * p<0.05, ** p<0.01, *** p < 0.001, **** p<0.0001, ns: no significant difference. Precise P values are provided in Raw Data.

Firstly, we observed an increase in the vitamin C level in C. elegans when fed with HK-yfbR (Figure 4B), a high carbohydrate food compared to HK-E. coli (Figure 3B and Figure 1-figure supplement 1E). However, the VC level in bacteria is the same (Figure 4-figure supplement 1A). The VC level also increased when D-glucose (Figure 4-figure supplement 1B) or D-glucuronate (D-GlcA) was added to HK-E. coli (Figure 4C), which was abolished by knocking down VC biosynthesis genes (Figure 4C and Figure 4-figure supplement 1B). This suggests that addition of sugar supplementation promotes VC synthesis in animals fed with HK-E. coli.

Secondly, we hypothesized that animals could overcome a low-quality food (HK-E. coli) environment by inhibiting the stress response through increasing vitamin C biosynthesis. We found that VC or D-glucuronate (D-GlcA) supplementation inhibits HK-E. coli induced UPRER (Figure 4D), immune response including irg-5/sysm-1 reporter expression (Figure 4E and Figure 4 – figure supplement 1C) and p-PMK-1 (Figure 4F and 4G), as well as food avoidance (Figure 4H).

Finally, we asked whether inhibition of stress response and avoidance by sugar supplementation depends on the vitamin C biosynthesis pathway. We found that suppression of HK-E. coli induced UPRER (Figure 4I), immune response (Figure 4J and Figure 4-figure supplement 1D) and food avoidance (Figure 4K) by D-GlcA/sugar supplementation was abolished in animals with RNAi of VC biosynthesis genes. Food selection behavior assays showed that D-GlcA/sugar supplementation increased the preference for heat-killed bacteria, which was also suppressed by knocking down VC biosynthesis genes (Figure 4-figure supplement 1E). However, VC supplementation still suppressed the UPRER (Figure 4I), immune response (Figure 4J and Figure 4-figure supplement 1D) and food avoidance (Figure 4K), and increased the food preference (Figure 4 – figure supplement 1E) in animals with or without RNAi of VC biosynthesis genes. This suggests that VC, as the final metabolite synthesized from D-glucose, is critical for evaluating low-quality food response in animals.

Together, these data indicate that the vitamin C biosynthesis pathway is critical for evaluating whether food is of higher quality and can be eaten by animals. It also suggests that animals could improve their VC levels to adapt to bad food environment.

Animals evaluate sugar and vitamin C through neuronal XBP-1 and PMK-1

As D-GlcA/sugar and VC supplementation suppressed HK-E. coli induced UPRER, immune response and food avoidance behavior, we investigated whether animals evaluate sugar and VC through XBP and PMK-1 dependent pathways. We performed a food selection behavior assay by adding D-Glc, D-GlcA or VC to the NGM, E. coli and HK-E. coli (Figure 5A). The food selection behavior assays revealed that supplementation with D-Glc, D-GlcA, or VC inhibits the animals’ choice of sugar or VC on E. coli-OP50 feeding conditions (Figure 5-figure supplement 1A). This suggests that supplementation with D-Glc, D-GlcA, or VC may alter the metabolites of live bacteria, leading to avoidance by the animals. There was no preference observed on NGM plate (no food condition) supplementation with D-Glc and VC (Figure 5-figure supplement 1B), indicating that the intake of sugar or VC alone does not influence animal preference. However, alone D-GlcA could influence worm physiology which induces preference change (Figure 5-figure supplement 1B). Interestingly, D-Glc and D-GlcA (Figure 5B and 5C) or VC (Figure 5D) supplementation increased the preference for heat-killed bacteria, which was suppressed in xbp-1 or pmk-1 mutant animals. However, this preference was also rescued in xbp-1 or pmk-1 mutant animals by expressing XBP-1 or PMK-1 in neurons rather than intestine (Figure 5B-D), indicating that neuronal XBP-1 and PMK-1 are critical for physiological food elevation system for monitoring the level of sugar and VC under low-quality food condition.

Animals evaluate sugar and vitamin C through neuronal XBP-1 and PMK-1

(A) Schematic method of the food selection assay. Food (red) and food with carbohydrate (D-Glc, D-GlcA, or VC) supplementation (green) was placed on indicated position. Synchronized L1 worms were then place in plate. After 16-17hs, the selection index was calculated.

(B-D) Food selection assay showing that xbp-1 or pmk-1 mutation eliminated the preference of HK-E. coli with D-Glc (B), D-GlcA (C) or Vitamin C (D) supplementation, which was rescued in xbp-1 or pmk-1 mutant animals by expressing XBP-1 or PMK-1 in neurons rather than intestine.

Data are represented as mean ±SD from five independent experiments, 68-647 animals/assay (B).

Data are represented as mean ±SD from six independent experiments, 83-701 animals/assay (C).

Data are represented as mean ±SD from six independent experiments, 67-1035 animals/assay (D).

For all panels, Scale bar shows on indicated figures, 50 μm. * p<0.05, ** p<0.01, *** p < 0.001, **** p<0.0001, ns: no significant difference. Precise P values are provided in Raw Data.

Discussion

To better survive, animals must evolve a system to recognize and evaluate the quality of their food. This includes the sensory neuron evaluation system for immediate response and feeding decision 2, as well as physiological food evaluation system for chronic response to ingested food. In our previous study, we discovered that the TORC1-ELT-2 pathway, acting as master regulators in intestine, evaluates vitamin B2 deficiency in low-quality food (HK-E. coli) and regulates gut digestive activity to impact animal’s food behavior 24. To further identified the mechanism by which animals evaluate low-quality food (HK-E. coli), we performed metabolomics and transcriptomics analyses to identify specific nutrition deficiencies in low-quality food and the cellular response pathways that are involved in food evaluation pathway. This study identified a physiological food evaluation mechanism by which animals recognize food quality through UPRER (IRE-1/XBP-1) – Innate immunity (PMK-1/p38 MAPK) regulated cellular stress response program in neurons that dictates food avoidance and selection behaviors (Figure 6).

Schematic model of physiological food evaluation system in evaluating/sensing sugar and vitamin C through UPRER (IRE-1/XBP-1) – Innate immunity (PMK-1/p38 MAPK) axis.

Vitamin C level is low in animals fed low sugar food, HK-E. coli. Sugar and Vitamin C deficiency activate cellular UPRER and immune response, which promote animals to leave low-quality food and seek better food for survival.

This cellular stress regulated physiological food evaluation system depends UPRER (IRE-1/XBP-1) – Innate immunity (PMK-1/p38 MAPK) axis in neuron.

One of the cellular stress response pathways, the Unfolded Protein Response (UPRER), is activated by various stresses, including infection and nutrition deficiency, which disrupt the homeostasis of the endoplasmic reticulum (ER)31. The activation of UPRER, specifically in the nervous system, has been shown to promote changes in feeding and foraging behavior47. The p38 PMK-1 pathway is also crucial for regulating the expression of secreted innate immune effectors and is essential for survival during infection48. Therefore, these two pathways play a critical role in ensuring animals’ survival in changing environments. However, it is still unclear whether UPRER and innate immunity evaluate food quality under physiological conditions. Our study provides evidence that low-quality food (HK-E. coli) activates both UPRER and p-PMK-1, leading to animals leaving the low-quality food. Previously finding have shown that inhibition of ER function promotes C. elegans avoid to the toxic food, which employs the MLK-1/MEK-1/KGB-1 pathway49. Notably, HK-E. coli induced avoidance behavior is independent of the KGB-1 pathway (Figure 5-figure supplement 1C). Additionally, our study reveals that neuronal UPRER and PMK-1 are essential for evaluating low-quality food, suggesting that the nervous system plays a critical role in assessing food quality.

Previous studies have shown that XBP1 deficiency in intestinal epithelial cells leads to IRE1a hyperactivation and increased JNK phosphorylation in the epithelial compartment in vivo50. The IRE1-XBP1 axis has been identified as a critical protective branch of the Unfolded Protein Response (UPR) induced secondary to an innate immune response in the presence of P. aeruginosa 18, 51. The p38 MAPK has also been shown to directly act on the phosphorylation of IRE-1 to promote the stress response52, 53. Interestingly, IRE-1 has been found to confer cold resistance independently of XBP-1 by activating JNK-1 MAPK49. In contrast, our study reveals a new mechanism where the UPRER (IRE-1/XBP-1) positively regulates Innate Immunity (PMK-1/p38 MAPK) under HK-E. coli food conditions, establishing a novel physiological food evaluation system that activates the cellular stress response program.

A previous study has shown that activating innate immunity (PMK-1 MAPK) leads to a reduction in translation54. Our own previous research has also demonstrated that PMK-1 activation causes a shutdown of food digestion in animals35, likely to reduce protein translation and cellular metabolism. To investigate this further, we measured the translation level of animals fed with HK-E. coli and found that total translation ability is significantly reduced in these animals (Figure 5-figure supplement 1D). This finding suggests that activating innate immunity (PMK-1 MAPK) may serve as a mechanism to slow down translation progress, thereby alleviating the pressure on the unfolded protein response (UPR) and preventing excessive UPRER activation.

Vitamin C (VC) is an important physiological antioxidant and a cofactor for a family of biosynthetic and gene regulatory monooxygenase and dioxygenase enzymes. It is also required for the biosynthesis of collagen, L-carnitine, and certain neurotransmitters 22, 23. Meanwhile, VC helps animals to protect neuron41, 42, defend excessive immune 43, 44, and inhibit inflammatory and ER stress45, 46 in order to better survive. The synthesis of vitamin C (VC) occurs through the glucuronate pathway, utilizing D-glucose as a precursor 39, 40 (Figure 4A). This led us to investigate whether the vitamin C biosynthesis pathway is involved in evaluating low-quality food by using D-glucose. In this study, we found that animals feeding live E. coli, which should produce more VC, exhibit higher glucose levels. However, our results show that animals maintain similar VC levels when fed ideal food (live E. coli) compared to low-quality food (HK-E. coli) (Figure 4B), suggesting that animals do not stimulate VC biosynthesis under favorable food conditions. In contrast, when animals are fed low-quality food (HK-OP50), we found that supplementing D-GlcA in HK-E. coli or E. coli-yfbR mutation can improve VC levels and inhibit UPRER-immunity (Figure 4C). These data indicate that glucose boosts the animal’s ability to adapt to unfavorable food environments by increasing VC levels, but not in favorable food conditions.

Unlike the sensory neuron evaluation system, which permits rapid feeding decisions through smell and taste, the cellular stress response as physiological food evaluation system describe here requires a slow and multi-step signal transduction process after the ingestion of food. The disruption of cellular homeostasis by ingested of low-quality or toxic food can activate stress response mechanisms that both increase the cellular ability to withstand and adapt to this disruption of homeostasis and promote behavioral strategies to avoid these conditions and lessen their impact on the organism. These cellular stress response mechanisms include heat shock response, unfolded protein response, oxidative stress response 55. Therefore, this slow physiological food evaluation system is an evolutionary adaptation mechanism for detecting nutrition deficiencies in food that was not detected by quick sensory nervous system.

One limitation of our study is the lack of explanation for why HK-E. coli activates UPRER and immunity. We hypothesized that when heat-killed, HK-E. coli may lack or contain altered levels of certain metabolites that either activate or inhibit UPRER and immunity, respectively. Additionally, we speculated that E. coli mutants killed by heat may lack metabolites that activate UPRER and immunity, or conversely, have increased levels of metabolites that inhibit these pathways. Fortunately, our investigation led to the discovery of the E. coli mutant yfbR, which inhibits UPRER and immunity by increasing carbohydrates that aid in resisting these stress pathways. Moving forward, we intend to further explore the intricate relationship between HK-E. coli and UPRER-immunity. This will be a key focus of our future research efforts.

Collectively, this study uncovers the unexpected function of UPRER (IRE-1/XBP-1) – Innate immunity (PMK-1/p38 MAPK) as a physiological food evaluation system for evaluating and sensing food quality in animals. It also highlights the utility of the HK-E. coli (low-quality food) – C. elegans interaction as a means to dissect the mechanism of food evaluation system in assessing food. Most importantly, it reveals that animals are capable of altering their nutrient (Vitamin C) levels through in vivo synthesis or food intake to adapt to a poor food environment when better food choices are not available.

Author Contributions

P. L designed, performed experiments, analyzed data. X. L constructed all transgenic animals. B.Q. designed research, supervised this study, and wrote the paper with inputs from P. L.

Supporting information